
\documentclass[preprint,12pt]{elsarticle}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{enumitem}
%\usepackage{thmtools}
\usepackage{hyperref}
\hypersetup{hypertexnames=false,hidelinks}
\usepackage[nameinlink,noabbrev]{cleveref}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{xcolor}


% ---- Theorem environments ----
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{claim}[theorem]{Claim}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% ---- Macros ----
\newcommand{\cond}{\operatorname{cond}}
\newcommand{\KL}{D_{\operatorname{KL}}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\Cov}{\operatorname{Cov}}
\newcommand{\tr}{\operatorname{Tr}}
\newcommand{\diag}{\operatorname{diag}}

\newcommand{\Range}{\operatorname{Range}}
\newcommand{\Span}{\operatorname{span}}
\newcommand{\id}{\mathrm{id}}
% \newcommand{\TODO}[1]{\textcolor{red}{[TODO: #1]}} % removed for submission

% ---- Cleveref names ----
\crefname{theorem}{theorem}{theorems}
\Crefname{theorem}{Theorem}{Theorems}
\crefname{proposition}{proposition}{propositions}
\Crefname{proposition}{Proposition}{Propositions}
\crefname{lemma}{lemma}{lemmas}
\Crefname{lemma}{Lemma}{Lemmas}
\crefname{corollary}{corollary}{corollaries}
\Crefname{corollary}{Corollary}{Corollaries}
\crefname{claim}{claim}{claims}
\Crefname{claim}{Claim}{Claims}
\crefname{definition}{definition}{definitions}
\Crefname{definition}{Definition}{Definitions}
\crefname{assumption}{assumption}{assumptions}
\Crefname{assumption}{Assumption}{Assumptions}
\crefname{example}{example}{examples}
\Crefname{example}{Example}{Examples}
\crefname{remark}{remark}{remarks}
\Crefname{remark}{Remark}{Remarks}
\crefname{equation}{Eq.}{Eqs.}
\Crefname{equation}{Equation}{Equations}
\crefname{figure}{fig.}{figs.}
\Crefname{figure}{Fig.}{Figs.}
\crefname{table}{table}{tables}
\Crefname{table}{Table}{Tables}
\crefname{section}{Sec.}{Secs.}
\Crefname{section}{Section}{Sections}

\journal{Journal of Functional Analysis}
\biboptions{numbers,sort&compress}

\begin{document}

\begin{frontmatter}

\title{Shorted Schur Complements for Trace-Class Covariance Blocks and Singular Gaussian Conditioning}

\author{Anonymous Author(s)}

    % ============================================================
    %  ABSTRACT
    % ============================================================
\begin{abstract}
        Schur complementation for compact trace-class Hilbert covariance blocks
        on a separable Hilbert space, under finite Cameron--Martin energy
        admissibility, is not an inversion problem but a shorting problem.  The surviving Schur
        object is Anderson--Trapp shorting, while the finite-dimensional formula
        $A-BD^{-1}B^*$ is only its invertible-block shadow.  In trace-class
        Hilbert problems, compact covariance blocks lack bounded ambient inverses,
        so shorting replaces direct inversion through a variational principle on
        the Cameron--Martin range.  For the affine finite-part normalization at rank one, the calibrated scaling is
        $\Sigma_{\lambda,kk}=\varepsilon_0/(2\lambda^2)$ and the shorting
        subtraction gives $S_{k,\lambda}=\frac{\varepsilon_0}{2\lambda^2}-R_\lambda$
        with $R_\lambda=O(\lambda^{-6})$.
        For Gaussian Radon laws, this yields the finite-part functional expansion
        \[
        \mathfrak{J}_\lambda(\mu_{\mathrm{obs}};\mu_\lambda)
        =
        \frac{b_k^2}{\varepsilon_0}+O(\lambda^{-4}),
        \]
        with a canonical evidence-dependent finite part at rank one.
    \end{abstract}

    \begin{keyword}
Block operators; Anderson--Trapp shorting; Schur complementation; Cameron--Martin range; trace-class covariance; Fredholm finite parts; Gaussian disintegration; Galerkin approximation
\end{keyword}

\end{frontmatter}

    % ============================================================
    %  \S1  INTRODUCTION
    % ============================================================
    \section{Introduction}
    \label{sec:intro}

    The starting point is a functional-analytic identification, not a
    probabilistic regularization.  In finite dimensions, the Schur complement
    of a positive block matrix is written as $A-BD^{-1}B^*$.  This formula is
    not the primitive operation: it is the invertible-block coordinate
    expression of Anderson--Trapp shorting
    \cite{krein1947selfadjoint,anderson1975shorted,douglas1966majorization,hassi2023complementation}.
    For compact trace-class covariance
    blocks on a separable Hilbert space, Anderson--Trapp shorting is the Schur
    object that survives when the free block $D$ has no bounded inverse on the
    ambient space
    \cite{bogachev1998gaussian,simon2005trace,daprato2014stochastic,jorgensentian2024hilbert}.  Singular
    conditioning therefore does not repair the Schur formula.  It exposes that
    Schur complementation was the finite-dimensional shadow of shorting all
    along.

    A systematic spectral obstruction result
    (Theorem~\ref{thm:triple_impossibility}) shows that direct inversion fails
    at three independent levels: algebraically there is no bounded inverse,
    numerically Galerkin truncations suffer spectral ill-conditioning
    $\kappa\sim N^{2}$, and asymptotically the regime may require
    $\lambda\gg N^{2}$, placing the true limit beyond the reach of any fixed
    truncation.  Classical Schur-complement and shorted-operator theory,
    originating in Krein's theory of non-negative extensions and developed by
    Anderson and Trapp, gives the underlying variational formula for positive
    block operators
    \cite{krein1947selfadjoint,anderson1975shorted,douglas1966majorization,hassi2023complementation}.
    Recent Schur-complement work on non-negative operators and complementable
    block operators gives a current operator-theoretic context for the same
    mechanism \cite{contino2021matrix,naikjohnson2024complementable}.
    In the covariance setting, its admissibility condition is exactly a
    Cameron--Martin range condition: cross-covariance vectors must have finite
    covariance energy rather than lie in the domain of a non-existent bounded
    inverse.  This paper uses that range condition to define the shorted
    residual as an energy projection, proves its $\lambda^{-6}$ scale, and
    shows that Galerkin approximation is Rayleigh--Ritz restriction of the same
    supremum to nested finite-dimensional test spaces.

    Gaussian conditioning enters after this operator identification is in
    place.  Singular coordinate conditioning reaches the Feldman--H\'{a}jek
    measure-class boundary: exact evidence is supported on a hyperplane of
    prior measure zero, while positive-noise Gaussian inverse-problem formulas may remain in
    the equivalent-measure regime
    \cite{cameron1944transformations,feldman1958equivalence,hajek1958property,bogachev1998gaussian,stuart2010inverse,carerelie2024covariance,carerelie2025mean}.
    The shorted covariance separates the finite residual from the diverging
    normalization terms.  Fredholm determinant identities and
    determinant-and-trace cancellation then extract a canonical finite part
    \cite{simon2005trace,gohberg2000traces}.

    Thus the paper's logical order is:
    shorting first, Gaussian disintegration second.  The stochastic terminology
    is an application layer.  The structural object is Anderson--Trapp shorting;
    the familiar Schur formula is its finite-dimensional invertible-block
    expression.

    In the rank-one case, the residual has the exact asymptotic form
    \[
    S_{k,\lambda}=\frac{\varepsilon_0}{2\lambda^2}-R_\lambda,
    \qquad
    R_\lambda=O(\lambda^{-6}),
    \]
    hence
    \[
    S_{k,\lambda}^{-1}=\frac{2}{\varepsilon_0}\lambda^2+O(\lambda^{-2}).
    \]
    These are the powers used in the finite-part cancellation: the target
    block contributes the $\lambda^2$ pole, while the shorted residual is too
    small to alter the finite part.

    The paper has four main contributions:
    \begin{enumerate}
        \item \textbf{Calibrated rank-one normalization.} We identify the
        rank-one block normalization that places the target variance in the
        affine finite-part regime and matches the balanced target scale
        $\Sigma_{\lambda,kk}=\varepsilon_0/(2\lambda^2)$.
        \item \textbf{Core operator shorting theorem.} We identify the formal
        inverse expression $A-BD^{-1}B^*$ as the invertible-block expression of
        Anderson--Trapp shorting, define the residual by the variational short
        on the Cameron--Martin range, and prove the $O(\lambda^{-6})$
        shorted-residual estimate.
        \item \textbf{Galerkin and Rayleigh--Ritz consistency.} Truncated
        Lyapunov covariances converge in trace norm and preserve the exact
        invariant under the finite-energy hypotheses.  Separately, fixed
        infinite-dimensional Rayleigh--Ritz restrictions satisfy
        $R_\lambda^{[n]}\uparrow R_\lambda$ and
        $S_{k,\lambda}^{[n]}\downarrow S_{k,\lambda}$.
        \item \textbf{Downstream Gaussian finite part.} For Gaussian
        Radon laws, disintegration applies the shorted covariance to the
        exact-evidence measure, while a rank-one Fredholm determinant identity
        cancels the trace anomaly and reduces the finite part to the invariant
        quadratic form.
    \end{enumerate}

    \subsection{Positioning in the literature}
    \label{sec:related}

    The paper is closest to four bodies of work.  First, it uses Schur
    complements, shorted operators, Douglas range inclusion, and modern
    complementability theory for block operators
    \cite{krein1947selfadjoint,anderson1975shorted,douglas1966majorization,tretter2008spectral,contino2018schur,friedrich2017generalized,contino2021matrix,naikjohnson2024complementable,hassi2023complementation}.
    Second, it uses the measure-class structure of Gaussian measures,
    especially the Cameron--Martin theorem and Feldman--H\'{a}jek dichotomy
    \cite{cameron1944transformations,feldman1958equivalence,hajek1958property,bogachev1998gaussian}.
    Third, it is adjacent to Bayesian inverse problems and Hilbert-space
    Gaussian conditioning, including shorted covariance descriptions, positive
    noise inverse-problem formulas, low-rank covariance approximations, and nonlinear
    observation maps
    \cite{owhadi2015conditioning,stuart2010inverse,steinwart2024conditioning,chen2024gaussian,carerelie2024covariance,carerelie2025mean,quang2022kl}.
    Fourth, the examples come from stochastic evolution equations and elliptic
    Gaussian fields, where dissipative sectorial generators produce compact
    trace-class stationary covariances naturally approximated by Galerkin
    truncation \cite{daprato2014stochastic,lord2014introduction,simon2005trace,henderson2022sobolev,jorgensentian2024hilbert}.

    \subsection{Organization of the paper}
    \label{sec:organization}

    \Cref{sec:setup} sets the operator model and canonical normalization.
    \Cref{sec:infdim} develops shorting, Galerkin/Rayleigh--Ritz approximation,
    and the GFF spectral obstruction to ambient Schur inversion.
    \Cref{sec:gaussian_zero} uses this shorted residual in Gaussian
    disintegration and the finite-part functional. \Cref{sec:spde} closes with an
    elliptic benchmark. \Cref{sec:conclusion} concludes. Appendix proofs are
    organized as: calibrated block identities and exact-resolvent formulas
    (\ref{app:block_identities}), bounded-inverse Schur shadows
    (\ref{app:shorting_residual}), Galerkin shorting consistency
    (\ref{app:galerkin_shorting}), Gaussian Fredholm analysis
    (\ref{app:gaussian_fredholm}), and admissibility geometry
    (\ref{app:admissibility_boundary}). Numerical diagnostics are collected
    in the supplementary material and are not proof dependencies.

    % ============================================================
    %  \S2 Operator Model and Canonical Normalization
    % ============================================================

    \section{Operator Model and Canonical Normalization}
    \label{sec:setup}

    \subsection{Rank-One Triangular Block Family}
    \label{sec:setup:evolution}

    We begin with a one-dimensional target block coupled into a free Hilbert
    block.  The block operator is triangular after the incoming target-coordinate
    blocks are removed; the parameter $\lambda>0$ controls only the target
    restoring rate and target diffusion scale.  This is the functional-analytic
    object whose covariance blocks will later be shorted.  A stochastic
    realization is the evolution on a separable Hilbert space $\mathcal{H}$
    \cite{daprato2014stochastic}:
    \begin{equation}
        \label{eq:stochastic_evolution}
        dX_t = \mathcal{A} X_t \, dt + \mathcal{D}^{1/2} dW_t
    \end{equation}
    where $\mathcal{A}$ is the drift operator and $\mathcal{D}$ is the trace-class diffusion operator. We decompose the Hilbert space as $\mathcal{H} = \mathcal{H}_k \oplus \mathcal{H}_I$, where $\mathcal{H}_k \coloneqq \mathbb{R}\phi_k$ is the one-dimensional target coordinate subspace, and $\mathcal{H}_I \coloneqq \mathcal{H}_k^\perp$ is the free bulk subspace. Let $\Pi_k \coloneqq \phi_k \otimes \phi_k$ and $\Pi_I \coloneqq I - \Pi_k$ be the corresponding orthogonal projections.

    We set the incoming target-coordinate blocks to zero and use $\lambda$ to
    define a rank-one triangular family around the target level $x_k^*$.

    \begin{definition}[Calibrated rank-one triangular family]
        \label{def:soft_do}
        The rank-one triangular family of strength $\lambda > 0$ at target
        coordinate $X_k$ modifies the drift and diffusion in
        \eqref{eq:stochastic_evolution} by
        \begin{equation}
            \label{eq:soft_do_sde}
            \mathcal{A}^{(\lambda)} \coloneqq \mathcal{A}_{\cond} - \lambda \Pi_k, \qquad
            \mathcal{D}^{(\lambda)} \coloneqq \mathcal{D}_{\cond} + d_\lambda \Pi_k
        \end{equation}
        where $\mathcal{A}_{\cond}$ is the block drift operator with incoming target-coordinate blocks set to zero ($\Pi_k \mathcal{A}_{\cond} = 0$), $\mathcal{D}_{\cond}$ has target-coordinate cross-noise set to zero, and $d_\lambda > 0$ is the local diffusion variance of the target block. In the SDE representation, this corresponds to:
        \begin{equation}
            \label{eq:soft_do_sde_general}
            dX_{k,t} = -\lambda (X_{k,t} - x_k^*) dt + \sqrt{d_\lambda}\, dW_{k,t}.
        \end{equation}
    \end{definition}

    \subsection{Canonical Normalization}
    \label{sec:setup:calibration}

    The block-decoupled structure under the triangular block family yields:
    \begin{equation}
        \label{eq:block_decoupled_drift_diffusion}
        \mathcal{A}^{(\lambda)}=
    \begin{pmatrix}-\lambda&0\\a_{Ik}&\mathcal{A}_{II}\end{pmatrix},\qquad
        \mathcal{D}^{(\lambda)}=
    \begin{pmatrix}d_\lambda&0\\0&D_{II}\end{pmatrix},
    \end{equation}
    where $a_{Ik} \coloneqq \mathcal{A}_{Ik} \phi_k \in \mathcal{H}_I$ represents the off-diagonal drift coupling.
    Let $\Sigma_\lambda$ denote the stationary covariance operator at strength
    $\lambda$, solving the Lyapunov equation:
    \begin{equation}
        \label{eq:stationary_lyapunov_regularized}
        \mathcal{A}^{(\lambda)}\Sigma_\lambda + \Sigma_\lambda(\mathcal{A}^{(\lambda)})^* + \mathcal{D}^{(\lambda)} = 0.
    \end{equation}
    The target block $kk$ decouples completely from the free coordinates, yielding the exact scalar equation:
    \begin{equation}
        \label{eq:target_scalar_lyapunov}
        -2\lambda\Sigma_{\lambda,kk} + d_\lambda = 0,
        \qquad
        \Sigma_{\lambda,kk} = \frac{d_\lambda}{2\lambda}.
    \end{equation}

    The normalization is calibrated relative to the affine finite-part criterion,
    not asserted as an operator inevitability.  In the Gaussian finite-part
    extension below, a constant forcing $b_k$ produces the deterministic target
    offset
    \begin{equation}
        \label{eq:target_offset_balancing}
        \Delta_\lambda \coloneqq m_{\lambda,k}-x_k^* = \frac{b_k}{\lambda}=O(\lambda^{-1}),
    \end{equation}
    so the squared deterministic displacement has scale
    $\Delta_\lambda^2=b_k^2/\lambda^2=O(\lambda^{-2})$.  The leading target
    contribution to the conditional finite-part functional is the
    signal-to-noise ratio
    \begin{equation}
        \label{eq:snr_balancing}
        \frac{\Delta_\lambda^2}{2\Sigma_{\lambda,kk}}
        =
        \frac{b_k^2/\lambda^2}{2\Sigma_{\lambda,kk}}.
    \end{equation}
    Consequently the balanced finite-part regime requires the target covariance
    to decay at the same order as the squared deterministic displacement.  If
    $\Sigma_{\lambda,kk}=O(\lambda^{-1})$, as for the slow family
    $d_\lambda=\varepsilon_0$, then \eqref{eq:snr_balancing} is
    $O(\lambda^{-1})$ and the finite target contribution collapses to zero.  If
    $\Sigma_{\lambda,kk}=O(\lambda^{-3})$, as for the fast family
    $d_\lambda=\varepsilon_0/\lambda^2$, then \eqref{eq:snr_balancing} is
    $O(\lambda)$ and the target contribution diverges.  The numerical supplement
    reports these slow and fast scaling diagnostics, but the conclusion already
    follows from the exact scalar identity.  A nonzero finite affine target limit
    therefore selects the critical covariance scale
    \begin{equation}
        \label{eq:critical_covariance_scale}
        \Sigma_{\lambda,kk}\asymp\lambda^{-2}.
    \end{equation}

    \begin{proposition}[Affine finite-part target scaling]
        \label{prop:exact_calibration}
        The unique target-diffusion scaling that makes the deterministic offset
        and target variance balance at a finite nonzero affine finite-part scale is
        \begin{equation}
            \label{eq:exact_variance_normalization}
            \Sigma_{\lambda,kk}\equiv \frac{\varepsilon_0}{2\lambda^2},
            \qquad \varepsilon_0>0.
        \end{equation}
        By the scalar Lyapunov identity \eqref{eq:target_scalar_lyapunov}, this is equivalent to
        \begin{equation}
            \label{eq:canonical_diffusion_scaling}
            d_\lambda = 2\lambda\Sigma_{\lambda,kk}
            = \frac{\varepsilon_0}{\lambda}
            \quad \text{for every } \lambda>0.
        \end{equation}
        With this canonical family,
        \begin{equation}
            \label{eq:canonical_balanced_ratio}
            \frac{\Delta_\lambda^2}{2\Sigma_{\lambda,kk}}
            = \frac{b_k^2}{\varepsilon_0},
        \end{equation}
        and the calibrated rank-one triangular family becomes
        \begin{equation}
            \label{eq:canonical_approximating_operators}
            dX_{k,t} = -\lambda (X_{k,t} - x_k^*) dt + \sqrt{\varepsilon_0 / \lambda}\, dW_{k,t}.
        \end{equation}
    \end{proposition}

    All subsequent results are stated for this canonical scaling family.

    \begin{definition}[Shorted Schur Energy Invariant]
        \label{def:schur_invariant}
        Let $S_{k,\lambda} \coloneqq \|X_k - P_{\mathcal{M}_I} X_k\|_{L^2}^2$ represent the shorted Schur projection residual of the target coordinate $X_k$ against the closed linear span $\mathcal{M}_I$ of the free coordinates $X_I$. In probabilistic language this is the linear minimum mean squared error residual, but the operator object is the Schur energy projection. The \emph{shorted Schur invariant} $m_2^{\mathrm{Schur}}$ is defined as:
        \begin{equation}
            \label{eq:m2_schur_definition}
            m_2^{\mathrm{Schur}} \coloneqq \lim_{\lambda\to\infty} \lambda^2 S_{k,\lambda}.
        \end{equation}
    \end{definition}

    \subsection{Operator Setup and Assumptions}
    \label{sec:setup:assumptions}

    We use one core assumption package for the shorting limit:

    \begin{assumption}[Core block-operator assumptions]
        \label{assum:core_analytical}
        \begin{enumerate}[label=(\roman*)]
            \item[(i)] \textbf{Trace-Class Lyapunov framework:} \label{assum:linear_second_moment}
            For each $\lambda>0$, the stationary covariance $\Sigma_\lambda$ is
            the unique positive injective trace-class solution to the trace-class
            Lyapunov equation
            \begin{equation}
                \label{eq:op_lyapunov}
                \mathcal{A}^{(\lambda)}\Sigma_\lambda + \Sigma_\lambda(\mathcal{A}^{(\lambda)})^* + \mathcal{D}^{(\lambda)} = 0.
            \end{equation}
            The free-block covariance under full decoupling is
            $Q_0\coloneqq\Sigma_{II}^{(0)}$, assumed compact, injective, trace-class,
            and positive.

            \item[(ii)] \textbf{Cameron--Martin admissibility of coupling:} \label{assum:cm_stability}
            The block-decoupled coefficients satisfy $\mathcal{A}_{kI}=\mathcal{D}_{kI}=\mathcal{D}_{Ik}=0$ and $\mathcal{A}_{kk}=-\lambda$.
            Let
            $H_{Q_0}=\Range(Q_0^{1/2})$ and
            $|u|_{Q_0}=\|Q_0^{-1/2}u\|_{\mathcal{H}_I}$. The leakage vector
            $a_{Ik}=\mathcal{A}_{Ik}\phi_k$ belongs to $H_{Q_0}$, and
            on $H_{Q_0}$ the restricted drift semigroup is exponentially stable:
            \begin{equation}
                \|e^{t\mathcal{A}_{II}}\|_{\mathcal{L}(H_{Q_0})} \le M e^{-\omega t},\qquad t\ge0,
            \end{equation}
            for some $M\ge1$ and $\omega>0$ independent of $\lambda$.
        \end{enumerate}
    \end{assumption}

    \begin{remark}[Finite-Energy Outgoing Covariance Coupling and Finite-Dimensional Calibration]
        \label{rem:cm_information_flow}
        The membership $a_{Ik}\in H_{Q_0}$ is an admissibility condition,
        not a consequence of the free-block Lyapunov equation alone: an arbitrary
        outgoing vector need not lie in
        $\Range((\Sigma_{II}^{(0)})^{1/2})$.  It says that the
        deterministic channel carrying target fluctuations into the free-block has
        finite covariance energy relative to the free-noise background.
  
        To understand this condition intuitively, consider a finite-dimensional setting where the free-block state space is $\mathcal{H}_I = \mathbb{R}^{d-1}$ and the free stationary covariance $Q_0 \coloneqq \Sigma_{II}^{(0)}$ is strictly positive-definite. Since $Q_0 \succ 0$, its square root $Q_0^{1/2}$ is invertible, and its range is the entire space $\mathbb{R}^{d-1}$. In this case, the Cameron--Martin space is the full space $H_{Q_0} = \Range(Q_0^{1/2}) = \mathbb{R}^{d-1}$, and the Cameron--Martin energy norm $|u|_{Q_0} = \|Q_0^{-1/2} u\|_2$ is equivalent to the standard Euclidean norm. Consequently, the admissibility condition $a_{Ik} \in H_{Q_0}$ is trivially satisfied for any outgoing vector $a_{Ik} \in \mathbb{R}^{d-1}$.
  
        However, if some direction in the free coordinates is noise-free under block decoupling (meaning $Q_0$ is positive semidefinite but singular), $H_{Q_0}$ becomes a proper subspace of $\mathbb{R}^{d-1}$. In this case, the condition $a_{Ik} \in H_{Q_0}$ requires that the leakage vector $a_{Ik}$ has no component in the null space of $Q_0$. Equivalently, it prevents the target coordinate from leaking fluctuations into a downstream coordinate that has zero background noise, which would otherwise result in infinite relative covariance energy.
  
        In infinite dimensions, $\Sigma_{II}^{(0)}$ is trace-class and compact, so its eigenvalues decay to zero, making $\Range((\Sigma_{II}^{(0)})^{1/2})$ a dense but strictly proper subspace of $\mathcal{H}_I$. The condition $a_{Ik} \in H_{Q_0}$ requires the off-diagonal drift coupling coefficients to decay sufficiently fast relative to the background noise eigenvalues to keep the relative energy finite.
  
        Under this assumption, invariance and exponential stability on $H_{Q_0}$ imply:
        \begin{equation}
            \begin{aligned}
                (\lambda I-\mathcal{A}_{II})^{-1}a_{Ik}
                &\in H_{Q_0},\\
                |(\lambda I-\mathcal{A}_{II})^{-1}a_{Ik}|_{Q_0}
                &\le \frac{M}{\lambda+\omega}|a_{Ik}|_{Q_0}.
            \end{aligned}
        \end{equation}
    \end{remark}

    \begin{remark}[Constant Dependency in Schur Energy Expansion]
        Consequently, the exact leakage cross-covariance
        $\Sigma_{\lambda,Ik}$ has finite Cameron--Martin energy norm and is
        $O(\lambda^{-3})$ in that norm.  This is a condition on the deterministic
        response and covariance energy; it does not assert that infinite-dimensional
        Gaussian sample paths lie in $H_{Q_0}$ almost surely.
    \end{remark}

    % ============================================================
    %  §3  INVERSE-FREE SCHUR COMPLEMENTATION
    % ============================================================
    \section{Inverse-Free Schur Complementation in Hilbert Space}
    \label{sec:infdim}

    \subsection{Variational Shorting and Energy Bounds}
    \label{sec:infdim:variational}

    In finite dimensions, the Schur complement of the bulk covariance block
    $\Sigma_{II}$ is written as
    $S_k \coloneqq \Sigma_{kk} - \Sigma_{kI}\Sigma_{II}^{-1}\Sigma_{Ik}$.
    Under stable and strictly positive diffusion, the bulk covariance
    $\Sigma_{II}$ is positive-definite and invertible, so this expression is a
    valid coordinate formula.  Its operator-theoretic content, however, is
    shorting: for a non-negative block operator, the Schur complement is the
    quadratic form of the Anderson--Trapp short to the target subspace, and the
    inverse expression appears only when the complementary block is invertible
    \cite{krein1947selfadjoint,anderson1975shorted}.  In infinite dimensions,
    the stationary covariance operator is trace-class and compact, so
    $\Sigma_{\lambda,II}$ lacks a bounded global inverse.  The finite-dimensional
    formula disappears, but the shorted operator remains well-defined through a
    variational supremum.  We therefore formulate the shorted residual directly
    in covariance-energy forms.

    Let
    \[
    Q_0
    =
    \int_0^\infty e^{r\mathcal A_{II}}D_{II}e^{r\mathcal A_{II}^*}\,dr
    \]
    denote the ambient stable-bulk background covariance from Assumption~\ref{assum:core_analytical}. We define the core Cameron--Martin space $H_{Q_0} \coloneqq \Range(Q_0^{1/2})$ and the core energy norm:
    \begin{equation}
        |u|_{Q_0} \coloneqq \|Q_0^{-1/2} u\|_{\mathcal{H}_I} \quad \text{for all } u \in H_{Q_0}.
    \end{equation}
    By Assumption~\ref{assum:cm_stability}, the off-diagonal drift coupling vector satisfies $a_{Ik} \in H_{Q_0}$, and the restricted drift generator $\mathcal{A}_{II}$ generates an exponentially stable semigroup on $H_{Q_0}$.

    Under this assumption, the exact cross-covariance resolvent formula (derived in \ref{app:block_identities}) satisfies:
    \begin{equation}
        \Sigma_{\lambda,Ik} = \frac{\varepsilon_0}{2\lambda^2} (\lambda I - \mathcal{A}_{II})^{-1} a_{Ik}.
    \end{equation}
    This resolvent response yields the Cameron--Martin energy bound on the cross-covariance block:
    \begin{equation}
        \label{eq:cross_norm_bound}
        |\Sigma_{\lambda,Ik}|_{Q_0} = O(\lambda^{-3}) \quad \text{as } \lambda \to \infty.
    \end{equation}

    To define the infinite-dimensional shorted residual, we introduce the variational Schur subtraction $R_\lambda$:
    \begin{equation}
        \label{eq:variational_schur_subtraction}
        R_\lambda \coloneqq \sup_{\langle u, \Sigma_{\lambda,II} u \rangle > 0} \frac{|\langle u, \Sigma_{\lambda,Ik} \rangle|^2}{\langle u, \Sigma_{\lambda,II} u \rangle}.
    \end{equation}
    Equivalently, if $\Sigma_\lambda$ is viewed as a non-negative block operator on $\mathbb{C}\phi_k\oplus\mathcal{H}_I$, then \eqref{eq:variational_schur_subtraction} is the one-dimensional Anderson--Trapp shorted-operator subtraction, namely
    \begin{equation}
        \label{eq:shorted_operator_scalar}
        \big\langle \phi_k,\mathcal{S}_{\mathbb{C}\phi_k}(\Sigma_\lambda)\phi_k\big\rangle
        =\Sigma_{\lambda,kk}-R_\lambda.
    \end{equation}
    The restriction to vectors with $\langle u,\Sigma_{\lambda,II}u\rangle>0$ simply removes null directions of the denominator; for a non-negative block operator, such directions do not change the supremum. Thus $R_\lambda$ is the energy removed by shorting to the target coordinate \cite{anderson1975shorted,owhadi2015conditioning,contino2021matrix,naikjohnson2024complementable}.

    Since the target noise channel and bulk noise channel are independent, the bulk state decomposes into independent background and target-driven components, meaning the bulk covariance dominates the background covariance ($\Sigma_{\lambda,II} = Q_0 + \operatorname{Cov}(Y_I) \succeq Q_0$). Douglas range inclusion then guarantees that the variational subtraction is bounded:
    \begin{equation}
        0 \le R_\lambda \le |\Sigma_{\lambda,Ik}|_{Q_0}^2 = O(\lambda^{-6}).
    \end{equation}

    \subsection{Shorted Schur Residual Scaling}
    \label{sec:infdim:scaling}

    The main scaling theorem for the shorted Schur residual follows. The $L^2$ projection interpretation is secondary and follows from covariance geometry.

    \begin{theorem}[Hilbert Shorted Schur Invariant]
        \label{thm:inf_cancellation}
        Under Assumption~\ref{assum:core_analytical}, the $L^2$-projection residual satisfies:
        \begin{equation}
            \label{eq:main_schur_sharp}
            S_{k,\lambda} \coloneqq \|X_k - P_{\mathcal{M}_I} X_k\|_{L^2}^2 = \frac{\varepsilon_0}{2\lambda^2} - R_\lambda,
        \end{equation}
        where the remainder satisfies $0 \le R_\lambda = O(\lambda^{-6})$. In particular, the shorted Schur invariant is:
        \begin{equation}
            \label{eq:m2_schur_value}
            m_2^{\mathrm{Schur}} = \lim_{\lambda\to\infty} \lambda^2 S_{k,\lambda} = \frac{\varepsilon_0}{2},
        \end{equation}
        and the reciprocal projection-residual scaling satisfies:
        \begin{equation}
            \label{eq:schur_scaling_inverse}
            S_{k,\lambda}^{-1} = \frac{2}{\varepsilon_0}\lambda^2 + O(\lambda^{-2}).
        \end{equation}
        This result depends only on covariance and orthogonal projection; it holds for any linear stationary model satisfying these assumptions, without Gaussianity.
    \end{theorem}
    \begin{proof}
        The block Lyapunov identities for the triangular family give
        \[
            \Sigma_{\lambda,kk}=\frac{\varepsilon_0}{2\lambda^2},
            \qquad
            \Sigma_{\lambda,Ik}
            =
            \frac{\varepsilon_0}{2\lambda^2}
            (\lambda I-\mathcal{A}_{II})^{-1}a_{Ik}.
        \]
        Since $a_{Ik}\in H_{Q_0}$ and the restricted resolvent is uniformly
        stable on $H_{Q_0}$,
        \[
            |\Sigma_{\lambda,Ik}|_{Q_0}
            \le
            \frac{\varepsilon_0 M}{2\lambda^2(\lambda+\omega)}
            |a_{Ik}|_{Q_0}
            =
            O(\lambda^{-3}).
        \]
        The independent stationary-convolution decomposition gives
        $\Sigma_{\lambda,II}\succeq Q_0$.  Douglas' range inclusion theorem
        therefore yields $H_{Q_0}\subseteq H_{\Sigma_{\lambda,II}}$ and
        $|v|_{\Sigma_{\lambda,II}}\le |v|_{Q_0}$ for $v\in H_{Q_0}$.

        The one-dimensional Anderson--Trapp shorting subtraction is the
        Rayleigh quotient
        \[
            R_\lambda
            =
            \sup_{\langle u,\Sigma_{\lambda,II}u\rangle>0}
            \frac{|\langle u,\Sigma_{\lambda,Ik}\rangle|^2}
                 {\langle u,\Sigma_{\lambda,II}u\rangle}
            =
            |\Sigma_{\lambda,Ik}|_{\Sigma_{\lambda,II}}^2.
        \]
        Hence
        \[
            0\le R_\lambda
            \le |\Sigma_{\lambda,Ik}|_{Q_0}^2
            =
            O(\lambda^{-6}).
        \]
        Shorting gives
        $S_{k,\lambda}=\Sigma_{\lambda,kk}-R_\lambda$, proving
        \eqref{eq:main_schur_sharp} and
        $m_2^{\mathrm{Schur}}=\varepsilon_0/2$.  Finally,
        \[
            S_{k,\lambda}^{-1}
            =
            \left(\frac{\varepsilon_0}{2\lambda^2}-O(\lambda^{-6})\right)^{-1}
            =
            \frac{2}{\varepsilon_0}\lambda^2+O(\lambda^{-2}).
        \]
        \ref{app:shorting_residual} gives the full block-operator
        derivation.
    \end{proof}

    \subsection{Galerkin Covariances and Rayleigh--Ritz Shorting}
    \label{sec:proofs:convergence}

    Two finite-dimensional procedures are needed, and they have different
    meanings.  First, the Galerkin Lyapunov covariance
    $\Sigma_{n,\lambda}$ is the covariance of the truncated triangular family;
    its shorted residual is denoted $R_{n,\lambda}$.  Second,
    $R_\lambda^{[n]}$ is the Rayleigh--Ritz restriction of the fixed
    infinite-dimensional variational shorting problem to a nested
    finite-dimensional test space.  Only the second construction carries the
    monotonicity statement.

    Let $\Pi_n: \mathcal{H} \to \mathcal{H}_n$ be finite-rank Galerkin projections preserving the target-bulk decomposition.  The covariance convergence statement is stated for admissible truncations used in the covariance-energy argument: on the bulk space, $\Pi_n^I$ truncates in an eigenbasis of $Q_0$, $Q_0^{(n)}=\Pi_n^I Q_0\Pi_n^I|_{\Pi_n^I\mathcal{H}_I}$, and $a_{Ik}^{(n)}=\Pi_n^I a_{Ik}$.  This choice preserves the Cameron--Martin energy uniformly:
    \begin{equation}
        \sup_n |a_{Ik}^{(n)}|_{Q_0^{(n)}}\le |a_{Ik}|_{Q_0}<\infty.
    \end{equation}

    Write $\mathcal{A}_n^{(\lambda)} \coloneqq \Pi_n \mathcal{A}^{(\lambda)}|_{\mathcal{H}_n}$, $\mathcal{D}_n^{(\lambda)} \coloneqq \Pi_n \mathcal{D}^{(\lambda)} \Pi_n$, and let $\Sigma_{n,\lambda}$ solve the truncated Lyapunov equation $\mathcal{A}_n^{(\lambda)} \Sigma_{n,\lambda} + \Sigma_{n,\lambda} (\mathcal{A}_n^{(\lambda)})^* + \mathcal{D}_n^{(\lambda)} = 0$.

    \begin{proposition}[Galerkin covariance and Rayleigh--Ritz shorting consistency]
        \label{prop:galerkin_consistency}
        Under Assumption~\ref{assum:core_analytical}, let $\Pi_n:\mathcal{H}\to\mathcal{H}_n$ be finite-rank orthogonal Galerkin projections preserving the target-bulk block decomposition:
        \begin{equation}
            \Pi_n\phi_k=\phi_k,\qquad \Pi_n\mathcal{H}_I\subset\mathcal{H}_I,\qquad \Pi_n\to I\ \text{strongly on }\mathcal{H}.
        \end{equation}
        Assume the truncated triangular operator family satisfies the uniform
        exponential stability and trace-class conditions stated in
        Lemma~\ref{lem:galerkin_trace_norm}, and assume either the uniform
        finite-energy condition
        \begin{equation}
            \sup_n |a_{Ik}^{(n)}|_{Q_0^{(n)}}<\infty,
        \end{equation}
        or, more concretely, the spectral truncation choice $a_{Ik}^{(n)}=\Pi_n^I a_{Ik}$ and $Q_0^{(n)}=\Pi_n^I Q_0\Pi_n^I$ in a common $Q_0$ eigenbasis.  Then the truncated covariance operators converge in trace norm:
        \begin{equation}
            \|\Sigma_{n,\lambda}-\Sigma_\lambda\|_{\mathcal{S}_1}\to0\quad\text{as }n\to\infty,
        \end{equation}
        and their shorted residuals satisfy
        \begin{equation}
            S_{k,\lambda}^{(n)}
            =
            \frac{\varepsilon_0}{2\lambda^2}-R_{n,\lambda},
            \qquad
            0\le R_{n,\lambda}\le C\lambda^{-6},
        \end{equation}
        with $C$ independent of $n$.  Thus the truncated Schur invariant is
        exact at every admissible level:
        \begin{equation}
            m_2^{\mathrm{Schur},(n)} \coloneqq \lim_{\lambda\to\infty} \lambda^2 S_{k,\lambda}^{(n)} = \frac{\varepsilon_0}{2} \quad \text{for all admissible } n.
        \end{equation}
        Separately, let $V_n\subset H_{\Sigma_{\lambda,II}}$ be nested
        finite-dimensional test spaces with dense union in the
        Cameron--Martin energy norm, and define
        \begin{equation}
            R_\lambda^{[n]}
            \coloneqq
            \sup_{\substack{u\in V_n\\ \langle u,\Sigma_{\lambda,II}u\rangle>0}}
            \frac{|\langle u,\Sigma_{\lambda,Ik}\rangle|^2}
                 {\langle u,\Sigma_{\lambda,II}u\rangle},
            \qquad
            S_{k,\lambda}^{[n]}
            \coloneqq
            \Sigma_{\lambda,kk}-R_\lambda^{[n]}.
        \end{equation}
        Then
        \begin{equation}
            R_\lambda^{[n]}\uparrow R_\lambda,
            \qquad
            S_{k,\lambda}^{[n]}\downarrow S_{k,\lambda}.
        \end{equation}
    \end{proposition}
    \begin{proof}
        See Lemmas~\ref{lem:galerkin_trace_norm},
        \ref{lem:galerkin_finite_energy_monotonicity}, and
        \ref{lem:galerkin_variational_shorting} for trace-norm covariance
        convergence, finite-energy control of spectral truncations, and the
        fixed-problem Rayleigh--Ritz statement.
    \end{proof}

    % ============================================================
    %  §3.4  SPECTRAL OBSTRUCTION TO AMBIENT SCHUR INVERSION
    % ============================================================
    \subsection{Spectral Obstruction to Ambient Schur Inversion in the GFF Regime}
    \label{sec:triple_impossibility}

    The variational shorting formula of \Cref{sec:infdim:variational} is the
    primitive operator object whose finite-dimensional invertible expression is
    the Schur formula.  The following result is deliberately stated in the GFF
    spectral regime used by the elliptic example \cite{sheffield2007gaussian,henderson2022sobolev}, rather than as a general
    compact-operator theorem.

\begin{theorem}[Spectral Obstruction to Ambient Schur Inversion in the GFF Regime]
        \label{thm:triple_impossibility}
        Let $Q_0\coloneqq\Sigma_{II}^{(0)}$ be the free-block covariance operator
        under full decoupling.  Assume the GFF spectral regime: in an orthonormal
        bulk eigenbasis,
        \begin{equation}
            Q_0 e_j=\sigma_j e_j,
            \qquad
            \sigma_j\asymp j^{-2},
        \end{equation}
        and the corresponding precision or drift eigenvalues satisfy
        \begin{equation}
            \mu_j\asymp j^2.
        \end{equation}
        Let $Q_0^{(N)}$ be the $N$-mode Galerkin truncation in this basis.  Then:
        \begin{enumerate}[label=(\roman*)]
            \item \label{item:algebraic_impossibility}
            \textbf{Algebraic impossibility.}
            In infinite dimensions, $Q_0$ is compact and injective with
            $0$ in its spectrum.  Hence $Q_0^{-1}$ is unbounded on the ambient
            Hilbert space, and the classical Schur expression
            $\Sigma_{KK}-\Sigma_{KI}\Sigma_{II}^{-1}\Sigma_{IK}$ is not an
            ambient bounded-operator formula.

            \item \label{item:numerical_impossibility}
            \textbf{Numerical impossibility.}
            The Galerkin covariance matrices have condition numbers
            \begin{equation}
                \label{eq:spectral_ill_conditioning}
                \kappa\bigl(Q_0^{(N)}\bigr)
                =
                \frac{\max_{1\le j\le N}\sigma_j}{\min_{1\le j\le N}\sigma_j}
                \asymp N^{2}.
            \end{equation}
            Direct inversion therefore amplifies errors at the same spectral rate
            as the inverse GFF precision scale.

            \item \label{item:asymptotic_impossibility}
            \textbf{Asymptotic impossibility.}
            The diagonal resolvent factor $(\lambda+\mu_j)^{-1}$ has a critical
            mode
            \begin{equation}
                \label{eq:gff_critical_mode}
                j^*(\lambda)\asymp \sqrt{\lambda}.
            \end{equation}
            Modes $j\ll j^*(\lambda)$ are in the clamp-dominated regime
            $(\lambda+\mu_j)^{-1}\asymp\lambda^{-1}$, while modes
            $j\gg j^*(\lambda)$ remain in the precision-dominated regime
            $(\lambda+\mu_j)^{-1}\asymp\mu_j^{-1}\asymp j^{-2}$.  Hence a fixed
            truncation dimension $N$ cannot uniformly represent the large-$\lambda$
            transition unless it resolves the moving scale $j^*(\lambda)$.
        \end{enumerate}

        The consistent resolution is the variational shorting operator
        (\Cref{sec:infdim:variational}), the operator object whose invertible-block
        coordinate expression is the Schur formula.  It is well-defined without
        bounded ambient inverses and its Galerkin approximation is a nested
        Rayleigh--Ritz restriction of the same variational quantity.
    \end{theorem}

    \begin{proof}
        For \ref{item:algebraic_impossibility}, compactness of $Q_0$ follows from
        $\sigma_j\to0$.  Since the Hilbert space is infinite-dimensional and
        $Q_0$ is injective with eigenvalues tending to zero, $0$ belongs to the
        spectrum but is not an eigenvalue.  Thus the inverse exists only as an
        unbounded operator on its range or as a Cameron--Martin form, not as a
        bounded operator on the ambient Hilbert space.

        For \ref{item:numerical_impossibility}, the assumptions
        $\sigma_j\asymp j^{-2}$ give constants $c,C>0$ such that
        $cj^{-2}\le\sigma_j\le Cj^{-2}$.  Hence
        $\max_{1\le j\le N}\sigma_j\asymp \sigma_1\asymp1$ and
        $\min_{1\le j\le N}\sigma_j\asymp\sigma_N\asymp N^{-2}$, proving
        \eqref{eq:spectral_ill_conditioning}.

        For \ref{item:asymptotic_impossibility}, the transition occurs where
        $\lambda$ and $\mu_j$ have comparable size.  Since $\mu_j\asymp j^2$,
        the relation $\mu_j\asymp\lambda$ is equivalent to
        $j\asymp\sqrt{\lambda}$, which proves \eqref{eq:gff_critical_mode}.  The
        two displayed resolvent regimes follow by comparing $\lambda$ and
        $\mu_j$ on either side of this scale.
    \end{proof}

    % ============================================================
    %  §4  FREDHOLM TRACE ANOMALIES AND GAUSSIAN FINITE PARTS
    % ============================================================
    \section{Fredholm Trace Anomalies and Gaussian Finite Parts}
    \label{sec:gaussian_zero}

    This section applies the shorted-operator construction to Gaussian Radon
    laws.  Disintegration supplies the exact-evidence measure; shorting supplies
    the conditional covariance object to which the Gaussian formulas can be
    applied.  The Fredholm finite part is then canonical: the determinant term
    and the trace-class perturbation generated by the shorted covariance defect
    cancel at the operator-spectral level.

    \subsection{Radon Gaussian Application and Disintegration}
    \label{sec:gaussian:setup}

    \begin{assumption}[Gaussian Application and Affine Mean Channel]
        \label{assum:gaussian_clamp}
        In addition to Assumption~\ref{assum:core_analytical}, let $\mu_\lambda = \mathcal{N}(m_\lambda, \Sigma_\lambda)$ be a Gaussian Radon law on $\mathcal{H}$ with injective trace-class covariance. Let $\mu_{\mathrm{obs}}$ be the canonical Gaussian disintegration member on the evidence hyperplane $X_k = x_k^*$. Suppose that a fixed deterministic forcing $b \in \mathcal{H}$ gives the stationary mean equation:
        \begin{equation}
            \mathcal{A}_{\cond} m_\lambda + b - \lambda \Pi_k (m_\lambda - x_k^* \phi_k) = 0, \quad \Pi_k \mathcal{A}_{\cond} = 0.
        \end{equation}
        We define the target deterministic offset constant:
        \begin{equation}
            c_\lambda \coloneqq \lambda(m_{\lambda,k} - x_k^*) = b_k,
        \end{equation}
        where $m_{\lambda,k} \coloneqq \langle m_\lambda, \phi_k \rangle$ and $b_k \coloneqq \langle b, \phi_k \rangle$.
    \end{assumption}

    Under the Gaussian Radon law, the coordinate projection $\pi_k: \mathcal{H} \to \mathbb{R}$ is continuous, so the target variable $X_k$ is a well-defined Borel random variable. The Polish structure of the Hilbert space guarantees the existence of a regular conditional probability distribution. 
    By selecting the weakly continuous version of the conditional law at the single point $t = x_k^*$ (as detailed in \ref{app:gaussian_fredholm}), we obtain the unique pointwise disintegration member $\mu_{\mathrm{obs}}$. Under this measure, the target coordinate satisfies the exact target zeros:
    \begin{equation}
        \label{eq:exact_target_zeros}
        m_{\text{obs},k} - x_k^* = 0, \quad \Var_{\text{obs}}(X_k) = 0 \quad (\text{exact}).
    \end{equation}

    \subsection{Measurable Cameron--Martin Energy Projection}
    \label{sec:gaussian:predictor}

    Let $C_\lambda \coloneqq \Sigma_{\lambda,II}$ denote the Gaussian bulk covariance operator on $\mathcal{H}_I$. Because $C_\lambda$ is trace-class and compact, it lacks a bounded global inverse. We define the bulk Cameron--Martin space $H_{C_\lambda} \coloneqq \Range(C_\lambda^{1/2})$ and the corresponding energy norm $|v|_{C_\lambda}\coloneqq\|C_\lambda^{-1/2}v\|_{\mathcal{H}_I}$.

    The leakage cross-covariance vector satisfies $\Sigma_{\lambda,Ik} \in H_{C_\lambda}$ with energy norm $|\Sigma_{\lambda,Ik}|_{C_\lambda} = O(\lambda^{-3})$. This range inclusion allows us to define the energy projection $\mu_{k|I}(X_I) \coloneqq \E[X_k \mid X_I]$ as a measurable Wiener functional:
    \begin{equation}
        \mu_{k|I}(X_I) = m_{\lambda,k} + W_{\Sigma_{\lambda,Ik}}(X_I - m_{\lambda,I}),
    \end{equation}
    which is the unique linear $L^2$ predictor. The estimates in \ref{app:gaussian_fredholm} imply that, for all sufficiently large $\lambda$, the bulk marginals $\mu_{\text{obs},I}$ and $\mu_{\lambda,I}$ are equivalent and have finite bulk relative entropy:
    \begin{equation}
        \KL(\mu_{\text{obs},I} \parallel \mu_{\lambda,I}) = O(\lambda^{-4}) \to 0 \quad \text{as } \lambda \to \infty.
    \end{equation}

    \begin{lemma}[Rank-one Fredholm determinant cancellation]
        \label{lem:rank_one_fredholm_cancellation}
        Let $C_\lambda=\Sigma_{\lambda,II}$ and
        $z_\lambda=\Sigma_{\lambda,Ik}$.  For all sufficiently large $\lambda$,
        the relative bulk covariance of the disintegration member has the form
        \begin{equation}
            C_\lambda^{-1/2}C_\lambda^{x_k^*}C_\lambda^{-1/2}
            =
            I-\rho_\lambda e_\lambda\otimes e_\lambda,
            \qquad
            \rho_\lambda
            =
            \frac{|z_\lambda|_{C_\lambda}^2}
                 {\Sigma_{\lambda,kk}}
            =
            O(\lambda^{-4}),
        \end{equation}
        with $\|e_\lambda\|=1$ when $z_\lambda\ne0$ and $\rho_\lambda=0$
        otherwise.  Hence
        \begin{equation}
            \det_{\!F}(I-\rho_\lambda e_\lambda\otimes e_\lambda)
            =
            1-\rho_\lambda,
            \qquad
            \tr(\rho_\lambda e_\lambda\otimes e_\lambda)
            =
            \rho_\lambda,
        \end{equation}
        and the covariance log-determinant/trace contribution satisfies
        \begin{equation}
            -\log(1-\rho_\lambda)-\rho_\lambda
            =
            O(\rho_\lambda^2).
        \end{equation}
    \end{lemma}
    \begin{proof}
        This is the rank-one Fredholm determinant identity.  A rank-one
        operator $e_\lambda\otimes e_\lambda$ has eigenvalue $1$ on
        $\mathbb{C}e_\lambda$ and vanishes on $e_\lambda^\perp$, so the relative
        covariance has eigenvalues $1-\rho_\lambda$ and $1$.  The determinant
        and trace formulas follow from this spectrum, and Taylor expansion of
        $-\log(1-\rho)-\rho$ at $\rho=0$ gives $O(\rho^2)$.
    \end{proof}

    \subsection{Gaussian Fredholm Finite-Part Functional}
    \label{sec:gaussian:functional}

    Although the joint KL divergence diverges under exact conditioning, the
    shorted covariance isolates the operator finite part before it is interpreted
    as a Gaussian density residue.

    \begin{definition}[Gaussian Fredholm Finite-Part Functional]
        \label{def:unified_action}
        Whenever $S_{k,\lambda}>0$ and $\mu_{\text{obs},I}$ is equivalent to $\mu_{\lambda,I}$, we define the Gaussian Fredholm finite-part functional of the conditioned law:
        \begin{equation}
            \label{eq:unified_action}
            \begin{aligned}
                \mathfrak{J}_\lambda(\mu_{\mathrm{obs}}; \mu_\lambda) \coloneqq &\KL(\mu_{\mathrm{obs},I} \parallel \mu_{\lambda,I}) \\
                &+ \frac{1}{2} S_{k,\lambda}^{-1} \E_{\mu_{\mathrm{obs}}}\left[ \left( X_k - \mu_{k|I}(X_I) \right)^2 \right].
            \end{aligned}
        \end{equation}
        This is the finite part of the joint KL divergence after removing the normalization defect due to the Dirac measure; the remaining covariance contribution is controlled by Fredholm determinant and trace terms.
    \end{definition}

    \begin{proposition}[Mollified KL extraction of the finite part]
        \label{prop:mollified_kl_extraction_main}
        For $\delta>0$, define
        \[
            \nu_{\lambda,\delta}
            \coloneqq
            \mu_{\mathrm{obs},I}\otimes \mathcal{N}(x_k^*,\delta^2)
        \]
        under the product decomposition
        $\mathcal{H}=\mathcal{H}_I\times\mathcal{H}_k$, and set
        \[
            C_{\lambda,\delta}
            \coloneqq
            \frac12\left(\log\frac{S_{k,\lambda}}{\delta^2}-1\right).
        \]
        Then, for all sufficiently large $\lambda$,
        \begin{equation}
            \lim_{\delta\downarrow0}
            \left[
                \KL(\nu_{\lambda,\delta}\Vert\mu_\lambda)
                -C_{\lambda,\delta}
            \right]
            =
            \mathfrak{J}_\lambda(\mu_{\mathrm{obs}};\mu_\lambda).
        \end{equation}
    \end{proposition}
    \begin{proof}
        This is Proposition~\ref{prop:mollification_limit}.
    \end{proof}

    \paragraph{Weak scheme-independence}
    For an evidence level $t\in\mathbb{R}$, write $\mu_\lambda^t$ for the weakly continuous conditional law from \ref{app:gaussian_fredholm} and set
    \[
    \mathfrak{J}_\lambda(t)\coloneqq
    \mathfrak{J}_\lambda(\mu_\lambda^t;\mu_\lambda).
    \]
    For any two evidence levels $t_1,t_2$ for which the finite part is defined,
    \begin{equation}
        \label{eq:weak_scheme_independence}
        \mathfrak{J}_\lambda(t_2)-\mathfrak{J}_\lambda(t_1)
        =
        \frac{(t_2-m_{\lambda,k})^2-(t_1-m_{\lambda,k})^2}
        {2\Sigma_{\lambda,kk}}.
    \end{equation}
    In the calibrated scaling $\Sigma_{\lambda,kk}=\varepsilon_0/(2\lambda^2)$, if
    $b_i=\lambda(m_{\lambda,k}-t_i)$, this becomes
    \[
    \mathfrak{J}_\lambda(t_2)-\mathfrak{J}_\lambda(t_1)
    =
    \frac{b_2^2-b_1^2}{\varepsilon_0}.
    \]
    The full proof is in \ref{app:gaussian_fredholm}, and it also shows
    weak-independence from the mollifier: replacing the Gaussian mollifier in
    Proposition~\ref{prop:mollification_limit} by any centered approximate identity with
    finite second moment changes the finite part only by a kernel-dependent term
    independent of $t$.

    \subsection{Gaussian Main Theorem}
    \label{sec:gaussian:theorem}

    The main Fredholm regularization and cancellation theorems for the finite-part functional follow.

    \begin{theorem}[Gaussian Fredholm Finite-Part Regularization]
        \label{thm:finite_part_regularization}
        Under Assumption~\ref{assum:gaussian_clamp}, there exists $\lambda_0>0$ such that, for $\lambda\ge\lambda_0$, the Gaussian Fredholm finite-part functional under the conditioned law $\mu_{\mathrm{obs}}$ is well-defined and satisfies:
        \begin{equation}
            \label{eq:unified_regularization}
            \mathfrak{J}_\lambda(\mu_{\mathrm{obs}}; \mu_\lambda) = \frac{b_k^2}{\varepsilon_0} + O(\lambda^{-4}).
        \end{equation}
        In particular, the functional remains bounded at $O(1)$ and converges to:
        \begin{equation}
            \lim_{\lambda\to\infty} \mathfrak{J}_\lambda(\mu_{\mathrm{obs}}; \mu_\lambda) = \frac{b_k^2}{\varepsilon_0}.
        \end{equation}
    \end{theorem}
    \begin{proof}
        See \ref{app:gaussian_fredholm} for the full derivation.
    \end{proof}

    \begin{corollary}[Exact Decoupled-Centered Cancellation]
        \label{cor:exact_pole_zero}
        Under Assumption~\ref{assum:gaussian_clamp}, if the target coordinate is decoupled from the bulk ($a_{Ik} = 0$) and the evidence is centered at the stationary mean ($x_k^* = m_{\lambda,k}$), then:
        \begin{equation}
            \mathfrak{J}_\lambda(\mu_{\mathrm{obs}}; \mu_\lambda) = 0 \quad \text{for every } \lambda > 0.
        \end{equation}
    \end{corollary}
    \begin{proof}
        The decoupling implies $\Sigma_{\lambda,Ik} = 0$, so $\KL(\mu_{\text{obs},I} \parallel \mu_{\lambda,I}) = 0$. The centering makes the conditional residual zero under the conditioning, so both terms in \eqref{eq:unified_action} vanish.
    \end{proof}

    % ============================================================
    %  §5  ELLIPTIC GAUSSIAN FIELD EXAMPLE
    % ============================================================
    \section{Elliptic Gaussian Field Example}
    \label{sec:spde}

    A standard elliptic Gaussian field example shows that the abstract Hilbert-space assumptions are natural for covariance operators arising from stochastic evolution equations and elliptic Gaussian priors \cite{daprato2014stochastic,lord2014introduction,stuart2010inverse,henderson2022sobolev}.

    \subsection{Elliptic Gaussian field on \texorpdfstring{$[0,1]$}{[0,1]}}

    Let
    \[
    L=-\frac{d^2}{dx^2}+\kappa^2
    \]
    on $L^2(0,1)$ with Dirichlet boundary conditions and $\kappa>0$. Then $\mathcal{A}_0=-L$ with domain $H^2(0,1)\cap H_0^1(0,1)$ is dissipative, $L$ is positive self-adjoint and sectorial, and $e^{-tL}$ is an exponentially stable analytic contraction semigroup. The inverse covariance
    \[
    Q_0=L^{-1}
    \]
    is compact, positive, self-adjoint, and trace class. With eigenfunctions $e_j(x)=\sqrt{2}\sin(\pi jx)$, its eigenvalues are
    \[
    q_j=(\pi^2j^2+\kappa^2)^{-1}.
    \]
    Thus $Q_0$ is a typical infinite-dimensional covariance operator: it is compact and trace class, has no bounded inverse on the ambient space, and carries its inverse only as a Cameron--Martin energy form \cite{bogachev1998gaussian,simon2005trace,henderson2022sobolev}.

    \subsection{Cameron--Martin range and covariance energy}

    The Cameron--Martin space associated with $Q_0$ is
    \[
    H_{Q_0}=\Range(Q_0^{1/2})=H_0^1(0,1),
    \]
    with equivalent energy norm
    \[
    \|h\|_{Q_0}^2
    =\langle h,Lh\rangle_{L^2}
    =\int_0^1 \bigl(|h'(x)|^2+\kappa^2|h(x)|^2\bigr)\,dx.
    \]
    This identifies the variational Schur projection in a Sobolev-energy form, as in the standard covariance-energy representation of Gaussian Hilbert priors \cite{bogachev1998gaussian,stuart2010inverse,henderson2022sobolev}. The covariance inverse does not act as a bounded operator on $L^2(0,1)$; it acts only through the energy form on its Cameron--Martin domain.

    \subsection{Spectral sensor conditioning}

    Fix a target eigenmode $e_k$ and define the observed coordinate
    \[
    X_k=\langle X,e_k\rangle_{L^2}.
    \]
    The unperturbed target variance is
    \[
    \Sigma_{0,kk}=q_k=(\pi^2k^2+\kappa^2)^{-1}.
    \]
    A regularized restoring drift on this coordinate produces the exact normalization
    \[
    \Sigma_{\lambda,kk}=\frac{\varepsilon_0}{2\lambda^2},
    \qquad
    S_{k,\lambda}=\frac{\varepsilon_0}{2\lambda^2}-R_\lambda,
    \qquad R_\lambda=O(\lambda^{-6}),
    \]
    under the block assumptions of \Cref{sec:infdim}. Off-diagonal bounded perturbations of the drift operator generate the cross-covariance vector entering the variational Schur residual. The resulting subtraction is evaluated by the Cameron--Martin energy form above, not by a global inverse on $L^2(0,1)$.

    This example shows that the abstract theorem applies to a genuine Gaussian field. The residual is controlled by a Sobolev energy on $H_0^1(0,1)$ while the ambient covariance remains compact.

    % ============================================================
    %  \S7  CONCLUSION
    % ============================================================
    \section{Conclusion}
    \label{sec:conclusion}

    Inverse-free Schur complementation for compact trace-class covariance blocks
    is Anderson--Trapp shorting, not ambient inversion.  The formal Schur algebra
    $A-BD^{-1}B^*$ is only its finite-dimensional invertible-block shadow.
    When the free covariance block is compact and trace class, the shadow
    disappears but the short remains: the residual is defined by variational
    shorting on the Cameron--Martin range, and Galerkin convergence follows
    from nested variational suprema rather than algebraic inversion.  Gaussian
    conditioning then follows as a Radon-measure consequence:
    disintegration supplies the exact-evidence law, and Fredholm
    determinant/trace identities cancel the finite-rank anomalies, leaving the
    canonical finite residual.  The paper does not treat infinite-rank
    observation operators or non-Gaussian conditioning theory.

    \section*{Declaration of generative AI and AI-assisted technologies in the writing process}

    During the preparation of this work, the author(s) used Google DeepMind's Antigravity assistant in order to merge the LaTeX manuscripts, clean up cross-referencing tags, format mathematical notation, and edit the draft for style consistency. After using this tool/service, the author(s) reviewed and edited the content as needed and take(s) full responsibility for the content of the published article.

    % ============================================================
    %  REFERENCES
    % ============================================================
    \bibliographystyle{elsarticle-num}
    \bibliography{paper1_ref}

    % ============================================================
    %  APPENDIX
    % ============================================================
    \appendix

    \input{paper1_appendix}
\end{document}
