
\documentclass[11pt]{amsart}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{enumitem}
%\usepackage{thmtools}
\usepackage{hyperref}
\hypersetup{hypertexnames=false,hidelinks}
\usepackage[nameinlink,noabbrev]{cleveref}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{xcolor}


% ---- Theorem environments ----
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{claim}[theorem]{Claim}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% ---- Macros ----
\newcommand{\cond}{\operatorname{cond}}
\newcommand{\KL}{D_{\operatorname{KL}}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\Cov}{\operatorname{Cov}}
\newcommand{\tr}{\operatorname{Tr}}
\newcommand{\diag}{\operatorname{diag}}

\newcommand{\Range}{\operatorname{Range}}
\newcommand{\Span}{\operatorname{span}}
\newcommand{\id}{\mathrm{id}}
% \newcommand{\TODO}[1]{\textcolor{red}{[TODO: #1]}} % removed for submission

% ---- Cleveref names ----
\crefname{theorem}{theorem}{theorems}
\Crefname{theorem}{Theorem}{Theorems}
\crefname{proposition}{proposition}{propositions}
\Crefname{proposition}{Proposition}{Propositions}
\crefname{lemma}{lemma}{lemmas}
\Crefname{lemma}{Lemma}{Lemmas}
\crefname{corollary}{corollary}{corollaries}
\Crefname{corollary}{Corollary}{Corollaries}
\crefname{claim}{claim}{claims}
\Crefname{claim}{Claim}{Claims}
\crefname{definition}{definition}{definitions}
\Crefname{definition}{Definition}{Definitions}
\crefname{assumption}{assumption}{assumptions}
\Crefname{assumption}{Assumption}{Assumptions}
\crefname{example}{example}{examples}
\Crefname{example}{Example}{Examples}
\crefname{remark}{remark}{remarks}
\Crefname{remark}{Remark}{Remarks}
\crefname{equation}{Eq.}{Eqs.}
\Crefname{equation}{Equation}{Equations}
\crefname{figure}{fig.}{figs.}
\Crefname{figure}{Fig.}{Figs.}
\crefname{table}{table}{tables}
\Crefname{table}{Table}{Tables}
\crefname{section}{Sec.}{Secs.}
\Crefname{section}{Section}{Sections}

\title[Block Operator Decoupling in Hilbert Space]{Block Operator Decoupling in Hilbert Space: Inverse-Free Schur Complementation and Gaussian Finite Parts}

\author{Anonymous Author(s)}

\begin{document}

    % ============================================================
    %  ABSTRACT
    % ============================================================
    \begin{abstract}
        We study the block-operator obstruction behind Schur complementation
        when the free covariance block is compact and trace class.  In this
        regime the finite-dimensional expression
        $A-BD^{-1}B^*$ is not an operator formula, because $D$ has no bounded
        ambient inverse.  We replace the inverse by the Anderson--Trapp
        variational shorting formula on the Cameron--Martin range, derive the
        canonical $\lambda^{-2}$ target-block scale from block Lyapunov
        normalization, prove an $O(\lambda^{-6})$ shorted-residual estimate,
        and extend the construction to finite-rank targets by Rayleigh--Ritz
        restriction of the same variational supremum.  Gaussian conditioning is
        then an application of this operator algebra: Galerkin monotonicity
        follows from subspace nesting, and the renormalized finite part is
        reduced by Fredholm determinant and trace identities to the invariant
        quadratic form
        \[
        \frac12\delta_E^\top\widetilde\Sigma_{EE}^{-1}\delta_E .
        \]
    \end{abstract}

    \keywords{Block operators; Schur complementation; shorted operators; Cameron--Martin range; trace-class covariance; Fredholm determinants; trace anomalies; Gaussian disintegration; Galerkin approximation}

    \maketitle

    % ============================================================
    %  \S1  INTRODUCTION
    % ============================================================
    \section{Introduction}
    \label{sec:intro}

    The starting point is a functional-analytic obstruction, not a
    probabilistic one.  In finite dimensions, the Schur complement of a
    positive block matrix is written as $A-BD^{-1}B^*$.  For covariance
    operators on a separable Hilbert space, the free block $D$ is typically
    compact and trace class; it therefore has no bounded inverse on the
    ambient space \cite{bogachev1998gaussian,simon2005trace,daprato2014stochastic}.
    The finite-dimensional Schur formula is then no longer an operator
    identity.  This is the algebraic obstruction addressed here.

    The replacement is Anderson--Trapp shorting.  Classical Schur-complement
    and shorted-operator theory, originating in Krein's theory of non-negative
    extensions and developed systematically by Anderson and Trapp, gives a
    variational formula for positive block operators
    \cite{krein1947selfadjoint,anderson1975shorted,douglas1966majorization}.
    In the covariance setting, its admissibility condition is exactly a
    Cameron--Martin range condition: cross-covariance vectors must have finite
    covariance energy rather than lie in the domain of a non-existent bounded
    inverse.  This paper uses that range condition to define the Schur residual
    as an energy projection, proves its $\lambda^{-6}$ scale, and shows that
    Galerkin approximation is only Rayleigh--Ritz restriction of the same
    supremum to nested finite-dimensional test spaces.

    Gaussian conditioning enters after this block-operator decoupling is in
    place.  Singular coordinate conditioning reaches the Feldman--H\'{a}jek
    measure-class boundary: exact evidence is supported on a hyperplane of
    prior measure zero, while positive-noise Gaussian inverse-problem formulas may remain in
    the equivalent-measure regime
    \cite{feldman1958equivalence,hajek1958property,bogachev1998gaussian,stuart2010inverse,carerelie2024optimal,carerelie2025posterior}.
    The block algebra separates the finite shorted covariance residual from
    the diverging normalization terms.  Fredholm determinant identities and
    trace-anomaly cancellation then give a finite Gaussian Fredholm
    functional.

    Thus the paper's logical order is:
    block operator decoupling first, Gaussian disintegration second.  The
    stochastic terminology is an application layer.  The structural object is
    inverse-free Schur complementation for compact covariance blocks.

    In the rank-one case, the residual has the exact asymptotic form
    \[
    S_{k,\lambda}=\frac{\varepsilon_0}{2\lambda^2}-R_\lambda,
    \qquad
    R_\lambda=O(\lambda^{-6}),
    \]
    hence
    \[
    S_{k,\lambda}^{-1}=\frac{2}{\varepsilon_0}\lambda^2+O(\lambda^{-2}).
    \]
    Section~\ref{sec:finite_rank} then replaces the scalar target block by a
    finite matrix $\Sigma_{\lambda,EE}=\lambda^{-2}\widetilde\Sigma_{EE}$,
    proves $\|R_{\lambda,E}\|=O(\lambda^{-6})$, and obtains the normalized
    matrix coupling $\rho_{\lambda,E}=O(\lambda^{-4})$. These are the powers
    used in the finite-part cancellation.

    The paper has four main contributions:
    \begin{enumerate}
        \item \textbf{Block Lyapunov normalization.} We identify the operator
        mechanism that fixes the target block at the unique finite-part scale
        $\Sigma_{\lambda,kk}=\varepsilon_0/(2\lambda^2)$.
        \item \textbf{Inverse-free Schur complementation.} We replace the
        failed free-block inverse by the Anderson--Trapp variational shorting
        formula on the Cameron--Martin range, and prove the $O(\lambda^{-6})$
        shorted-residual estimate.
        \item \textbf{Variational Galerkin monotonicity.} We prove that
        finite-dimensional truncations converge by monotone expansion of the
        variational supremum.  Loewner monotonicity is a consequence of nested
        subspaces, not matrix inversion.
        \item \textbf{Fredholm trace-anomaly cancellation.} For Gaussian
        Radon laws, disintegration supplies the exact-evidence measure, while
        Weinstein--Aronszajn/Fredholm determinant identities cancel the trace
        anomalies and reduce the finite part to the invariant quadratic form.
    \end{enumerate}

    \subsection{Positioning in the literature}
    \label{sec:related}

    The paper is closest to four bodies of work.  First, it uses Schur
    complements, shorted operators, Douglas range inclusion, and modern
    complementability theory for block operators
    \cite{krein1947selfadjoint,anderson1975shorted,douglas1966majorization,tretter2008spectral,contino2018schur,friedrich2017generalized,hassi2023complementation}.
    Second, it uses the measure-class structure of Gaussian measures,
    especially the Cameron--Martin theorem and Feldman--H\'{a}jek dichotomy
    \cite{feldman1958equivalence,hajek1958property,bogachev1998gaussian}.
    Third, it is adjacent to Bayesian inverse problems and Hilbert-space
    Gaussian conditioning, including shorted covariance descriptions, positive
    noise inverse-problem formulas, low-rank covariance approximations, and nonlinear
    observation maps
    \cite{owhadi2015conditioning,stuart2010inverse,chen2024gaussian,steinwart2024conditioning,carerelie2024optimal,carerelie2025posterior}.
    Fourth, the examples come from stochastic evolution equations and elliptic
    Gaussian fields, where dissipative sectorial generators produce compact
    trace-class stationary covariances naturally approximated by Galerkin
    truncation \cite{daprato2014stochastic,lord2014introduction,simon2005trace}.

    \subsection{Organization of the paper}
    \label{sec:organization}

    \Cref{sec:setup} defines the block operator model and canonical
    normalization. \Cref{sec:infdim} establishes inverse-free Hilbert-space
    Schur complementation. \Cref{sec:gaussian_zero} applies the operator
    result to Radon Gaussian disintegration and Fredholm finite parts.
    \Cref{sec:spde} gives an elliptic Gaussian-field example.
    \Cref{sec:finite_rank} develops finite-rank shorting, operator-valued
    residuals, misalignment, and exact matrix/Fredholm cancellation.
    \Cref{sec:conclusion} concludes. Proofs are given in
    \Cref{app:block_cumulant_analysis,app:gaussian_lift}; numerical diagnostics
    and Galerkin verification are given in \Cref{app:experiments}.

    % ============================================================
    %  \S2 Operator Model and Canonical Normalization
    % ============================================================

    \section{Operator Model and Canonical Normalization}
    \label{sec:setup}

    \subsection{Operator Dynamics and Regularized Conditioning}
    \label{sec:setup:dynamics}

    We model the continuous-time dynamics as stochastic evolution on a separable Hilbert space $\mathcal{H}$ \cite{daprato2014stochastic}:
    \begin{equation}
        \label{eq:stochastic_evolution}
        dX_t = \mathcal{A} X_t \, dt + \mathcal{D}^{1/2} dW_t
    \end{equation}
    where $\mathcal{A}$ is the drift operator and $\mathcal{D}$ is the trace-class diffusion operator. We decompose the Hilbert space as $\mathcal{H} = \mathcal{H}_k \oplus \mathcal{H}_I$, where $\mathcal{H}_k \coloneqq \mathbb{R}\phi_k$ is the one-dimensional target coordinate subspace, and $\mathcal{H}_I \coloneqq \mathcal{H}_k^\perp$ is the free bulk subspace. Let $\Pi_k \coloneqq \phi_k \otimes \phi_k$ and $\Pi_I \coloneqq I - \Pi_k$ be the corresponding orthogonal projections.

    To approximate the exact point conditioning $X_k = x_k^*$, we set the incoming target-coordinate blocks to zero and apply a regularization parameter $\lambda > 0$.

    \begin{definition}[Calibrated Approximating Dynamics Family]
        \label{def:soft_do}
        The regularized conditioning operator of strength $\lambda > 0$ at coordinate $X_k$ acts on the operator dynamics \eqref{eq:stochastic_evolution} by modifying both the drift and the diffusion:
        \begin{equation}
            \label{eq:soft_do_sde}
            \mathcal{A}^{(\lambda)} \coloneqq \mathcal{A}_{\cond} - \lambda \Pi_k, \qquad
            \mathcal{D}^{(\lambda)} \coloneqq \mathcal{D}_{\cond} + d_\lambda \Pi_k
        \end{equation}
        where $\mathcal{A}_{\cond}$ is the block drift operator with incoming target-coordinate blocks set to zero ($\Pi_k \mathcal{A}_{\cond} = 0$), $\mathcal{D}_{\cond}$ has target-coordinate cross-noise set to zero, and $d_\lambda > 0$ is the local diffusion variance of the target coordinate at regularization strength $\lambda$. In the SDE, this corresponds to:
        \begin{equation}
            \label{eq:soft_do_sde_general}
            dX_{k,t} = -\lambda (X_{k,t} - x_k^*) dt + \sqrt{d_\lambda}\, dW_{k,t}.
        \end{equation}
    \end{definition}

    \subsection{Canonical Normalization}
    \label{sec:setup:calibration}

    The block-decoupled structure under the regularized dynamics yields:
    \begin{equation}
        \label{eq:block_decoupled_drift_diffusion}
        \mathcal{A}^{(\lambda)}=
    \begin{pmatrix}-\lambda&0\\a_{Ik}&\mathcal{A}_{II}\end{pmatrix},\qquad
        \mathcal{D}^{(\lambda)}=
    \begin{pmatrix}d_\lambda&0\\0&D_{II}\end{pmatrix},
    \end{equation}
    where $a_{Ik} \coloneqq \mathcal{A}_{Ik} \phi_k \in \mathcal{H}_I$ represents the off-diagonal drift coupling.
    Let $\Sigma_\lambda$ denote the stationary covariance operator under regularization strength $\lambda$, solving the Lyapunov equation:
    \begin{equation}
        \label{eq:stationary_lyapunov_regularized}
        \mathcal{A}^{(\lambda)}\Sigma_\lambda + \Sigma_\lambda(\mathcal{A}^{(\lambda)})^* + \mathcal{D}^{(\lambda)} = 0.
    \end{equation}
    The target block $kk$ decouples completely from the free coordinates, yielding the exact scalar equation:
    \begin{equation}
        \label{eq:target_scalar_lyapunov}
        -2\lambda\Sigma_{\lambda,kk} + d_\lambda = 0,
        \qquad
        \Sigma_{\lambda,kk} = \frac{d_\lambda}{2\lambda}.
    \end{equation}

    The normalization is fixed by canonical scaling, not by a free modeling convention.  In the Gaussian finite-part extension below, a constant forcing $b_k$ produces the deterministic target offset
    \begin{equation}
        \label{eq:target_offset_balancing}
        \Delta_\lambda \coloneqq m_{\lambda,k}-x_k^* = \frac{b_k}{\lambda}=O(\lambda^{-1}),
    \end{equation}
    so the squared deterministic displacement has the rigid scale $\Delta_\lambda^2=b_k^2/\lambda^2=O(\lambda^{-2})$.  The leading target contribution to the conditional finite-part functional is the signal-to-noise ratio
    \begin{equation}
        \label{eq:snr_balancing}
        \frac{\Delta_\lambda^2}{2\Sigma_{\lambda,kk}}
        =
        \frac{b_k^2/\lambda^2}{2\Sigma_{\lambda,kk}}.
    \end{equation}
    Consequently the target covariance must decay at the same order as the squared deterministic displacement.  If $\Sigma_{\lambda,kk}=O(\lambda^{-1})$, as for the slow family $d_\lambda=\varepsilon_0$, then \eqref{eq:snr_balancing} is $O(\lambda^{-1})$ and the finite target contribution collapses to zero.  If $\Sigma_{\lambda,kk}=O(\lambda^{-3})$, as for the fast family $d_\lambda=\varepsilon_0/\lambda^2$, then \eqref{eq:snr_balancing} is $O(\lambda)$ and the target contribution diverges.  These two failures are the slow and fast ablations reported in \Cref{app:experiments:ablation}.  A nonzero finite limit therefore forces the critical covariance scale
    \begin{equation}
        \label{eq:critical_covariance_scale}
        \Sigma_{\lambda,kk}\asymp\lambda^{-2}.
    \end{equation}

    \begin{proposition}[Canonical Scaling]
        \label{prop:exact_calibration}
        The unique target-diffusion scaling that makes the deterministic offset and target variance balance at a finite nonzero scale is
        \begin{equation}
            \label{eq:exact_variance_normalization}
            \Sigma_{\lambda,kk}\equiv \frac{\varepsilon_0}{2\lambda^2},
            \qquad \varepsilon_0>0.
        \end{equation}
        By the scalar Lyapunov identity \eqref{eq:target_scalar_lyapunov}, this is equivalent to
        \begin{equation}
            \label{eq:canonical_diffusion_scaling}
            d_\lambda = 2\lambda\Sigma_{\lambda,kk}
            = \frac{\varepsilon_0}{\lambda}
            \quad \text{for every } \lambda>0.
        \end{equation}
        With this canonical family,
        \begin{equation}
            \label{eq:canonical_balanced_ratio}
            \frac{\Delta_\lambda^2}{2\Sigma_{\lambda,kk}}
            = \frac{b_k^2}{\varepsilon_0},
        \end{equation}
        and the calibrated approximating dynamics become
        \begin{equation}
            \label{eq:canonical_approximating_dynamics}
            dX_{k,t} = -\lambda (X_{k,t} - x_k^*) dt + \sqrt{\varepsilon_0 / \lambda}\, dW_{k,t}.
        \end{equation}
    \end{proposition}

    All subsequent results are stated for this canonical scaling family.

    \begin{definition}[Shorted Schur Energy Invariant]
        \label{def:schur_invariant}
        Let $S_{k,\lambda} \coloneqq \|X_k - P_{\mathcal{M}_I} X_k\|_{L^2}^2$ represent the shorted Schur projection residual of the target coordinate $X_k$ against the closed linear span $\mathcal{M}_I$ of the free coordinates $X_I$. In probabilistic language this is the linear minimum mean squared error residual, but the operator object is the Schur energy projection. The \emph{shorted Schur invariant} $m_2^{\mathrm{Schur}}$ is defined as:
        \begin{equation}
            \label{eq:m2_schur_definition}
            m_2^{\mathrm{Schur}} \coloneqq \lim_{\lambda\to\infty} \lambda^2 S_{k,\lambda}.
        \end{equation}
    \end{definition}

    \subsection{Operator Setup and Assumptions}
    \label{sec:setup:assumptions}

    We state the operator-theoretic assumptions that make the limit passage independent of the Galerkin dimension.

    \begin{definition}[Drift Operator with Analytic Base]
        \label{def:drift_operator}
        A drift operator on a separable Hilbert space $\mathcal{H}$ is an operator of the form $\mathcal{A} = \mathcal{A}_0 + \mathcal{K}$, where:
        \begin{enumerate}
            \item[(i)] $\mathcal{A}_0: \operatorname{dom}(\mathcal{A}_0) \subset \mathcal{H} \to \mathcal{H}$ is closed, densely defined, and dissipative, and it generates a strongly continuous analytic contraction semigroup $(e^{t\mathcal{A}_0})_{t \ge 0}$ on $\mathcal{H}$ satisfying $\|e^{t\mathcal{A}_0}\| \le e^{-\alpha t}$ for some spectral gap $\alpha > 0$; equivalently, the base operator $-\mathcal{A}_0$ is sectorial of angle $<\pi/2$ in this exponentially stable contraction setting.
            \item[(ii)] $\mathcal{K}: \mathcal{H} \to \mathcal{H}$ is a bounded linear operator.
            \item[(iii)] The full operator $\mathcal{A} = \mathcal{A}_0 + \mathcal{K}$, with $\operatorname{dom}(\mathcal{A})=\operatorname{dom}(\mathcal{A}_0)$, generates a strongly continuous analytic semigroup $(e^{t\mathcal{A}})_{t \ge 0}$ by the bounded perturbation theorem, and this semigroup remains exponentially stable: $\|e^{t\mathcal{A}}\| \le M e^{-\beta t}$ for some $M \ge 1$, $\beta > 0$.
        \end{enumerate}
        In finite dimensions ($\mathcal{H} = \mathbb{R}^n$), this reduces to a decomposition of a Hurwitz drift matrix into an exponentially stable analytic base plus a bounded perturbation.
    \end{definition}

    \begin{definition}[Trace-Class Noise and Covariance]
        \label{def:trace_class}
        Let $\mathcal{S}_1(\mathcal{H})$ denote the Banach space of trace-class (nuclear) operators on $\mathcal{H}$ equipped with the trace norm $\|T\|_{\mathcal{S}_1} \coloneqq \tr(|T|) = \sum_{i=1}^\infty \langle \phi_i, (T^*T)^{1/2} \phi_i \rangle < \infty$.
        \begin{enumerate}
            \item[(i)] The diffusion operator $\mathcal{D}: \mathcal{H} \to \mathcal{H}$ is positive, self-adjoint, and belongs to $\mathcal{S}_1(\mathcal{H})$.
            \item[(ii)] Assuming the pair $(\mathcal{A}, \mathcal{D}^{1/2})$ is controllable, the stationary covariance operator $\Sigma: \mathcal{H} \to \mathcal{H}$ is the unique strictly positive, self-adjoint, trace-class solution to the operator Lyapunov equation:
            \begin{equation}
                \label{eq:op_lyapunov}
                \mathcal{A}\Sigma + \Sigma\mathcal{A}^* + \mathcal{D} = 0
            \end{equation}
            which is given by the convergent Bochner integral
            \[
            \Sigma = \int_0^\infty e^{t\mathcal{A}} \mathcal{D} e^{t\mathcal{A}^*} dt.
            \]
        \end{enumerate}
    \end{definition}

    \begin{assumption}[Core Analytical Assumptions for Singular Hyperplane Conditioning]
        This work relies on two core analytical assumptions regarding the target-coordinate block decoupling and the outgoing coupling:
        \begin{enumerate}
            \item[(i)] \textbf{Linear Second-Moment Channel:} \label{assum:linear_second_moment} A centered linear stationary process on $\mathcal{H}=\mathcal{H}_k\oplus \mathcal{H}_I$ has finite second moments and covariance solving \Cref{eq:op_lyapunov}. The block-decoupling constraint zeroes all incoming target drift and cross-noise, so that $\mathcal{A}_{kI}=\mathcal{D}_{kI}=\mathcal{D}_{Ik}=0$ and $\mathcal{A}_{kk}=-\lambda$. The target diffusion $\mathcal{D}_{kk}=\varepsilon_0/\lambda$ is the unique scaling determined by the canonical normalization (Proposition~\ref{prop:exact_calibration}). The free noise is independent of the local target noise. (Gaussianity is not assumed).
            \item[(ii)] \textbf{Cameron--Martin Stability of Off-Diagonal Drift Coupling:} \label{assum:cm_stability} Let $Q_0\coloneqq\Sigma_{II}^{(0)}$ denote the free-block covariance produced by the free noise under block decoupling, and let $H_{Q_0}\coloneqq\operatorname{Range}(Q_0^{1/2})$ with the energy norm $|u|_{Q_0}\coloneqq\|Q_0^{-1/2}u\|_{\mathcal{H}_I}$. The outgoing vector $a_{Ik}\coloneqq\mathcal{A}_{Ik}\phi_k$ belongs to $H_{Q_0}$, and $e^{t\mathcal{A}_{II}}$ restricts to a strongly continuous exponentially stable semigroup on $H_{Q_0}$:
            \begin{equation}
                \|e^{t\mathcal{A}_{II}}\|_{\mathcal{L}(H_{Q_0})} \le M e^{-\omega t},\qquad t\ge0,
            \end{equation}
            for constants $M\ge1$ and $\omega>0$ independent of $\lambda$.
        \end{enumerate}
        \vspace{1mm}
        \noindent\textit{Intuitive Explanation:} Part (i) states that the regularized dynamics act as a strong localized drift penalization on the target coordinate $X_k$ while isolating it from incoming influences, giving an operator implementation of the block-decoupling constraint. Part (ii) requires that the off-diagonal drift coupling $a_{Ik}$ from the target coordinate has finite energy relative to the background fluctuations of the free coordinates. This ensures that target fluctuations propagating downstream do not overload the variance of the infinite-dimensional system.
    \end{assumption}

    \begin{remark}[Finite-Energy Outgoing Covariance Coupling and Finite-Dimensional Calibration]
        \label{rem:cm_information_flow}
        The membership $a_{Ik}\in H_{Q_0}$ is an admissibility condition,
        not a consequence of the free-block Lyapunov equation alone: an arbitrary
        outgoing vector need not lie in
        $\operatorname{Range}((\Sigma_{II}^{(0)})^{1/2})$.  It says that the
        deterministic channel carrying target fluctuations into the free-block has
        finite covariance energy relative to the free-noise background.
  
        To understand this condition intuitively, consider a finite-dimensional setting where the free-block state space is $\mathcal{H}_I = \mathbb{R}^{d-1}$ and the free stationary covariance $Q_0 \coloneqq \Sigma_{II}^{(0)}$ is strictly positive-definite. Since $Q_0 \succ 0$, its square root $Q_0^{1/2}$ is invertible, and its range is the entire space $\mathbb{R}^{d-1}$. In this case, the Cameron--Martin space is the full space $H_{Q_0} = \operatorname{Range}(Q_0^{1/2}) = \mathbb{R}^{d-1}$, and the Cameron--Martin energy norm $|u|_{Q_0} = \|Q_0^{-1/2} u\|_2$ is equivalent to the standard Euclidean norm. Consequently, the admissibility condition $a_{Ik} \in H_{Q_0}$ is trivially satisfied for any outgoing vector $a_{Ik} \in \mathbb{R}^{d-1}$.
  
        However, if some direction in the free coordinates is noise-free under block decoupling (meaning $Q_0$ is positive semidefinite but singular), $H_{Q_0}$ becomes a proper subspace of $\mathbb{R}^{d-1}$. In this case, the condition $a_{Ik} \in H_{Q_0}$ requires that the leakage vector $a_{Ik}$ has no component in the null space of $Q_0$. Equivalently, it prevents the target coordinate from leaking fluctuations into a downstream coordinate that has zero background noise, which would otherwise result in infinite relative covariance energy.
  
        In infinite dimensions, $\Sigma_{II}^{(0)}$ is trace-class and compact, so its eigenvalues decay to zero, making $\operatorname{Range}((\Sigma_{II}^{(0)})^{1/2})$ a dense but strictly proper subspace of $\mathcal{H}_I$. The condition $a_{Ik} \in H_{Q_0}$ requires the off-diagonal drift coupling coefficients to decay sufficiently fast relative to the background noise eigenvalues to keep the relative energy finite.
  
        Under this assumption, invariance and exponential stability on $H_{Q_0}$ imply:
        \begin{equation}
            \begin{aligned}
                (\lambda I-\mathcal{A}_{II})^{-1}a_{Ik}
                &\in H_{Q_0},\\
                |(\lambda I-\mathcal{A}_{II})^{-1}a_{Ik}|_{Q_0}
                &\le \frac{M}{\lambda+\omega}|a_{Ik}|_{Q_0}.
            \end{aligned}
        \end{equation}
    \end{remark}

    \begin{remark}[Constant Dependency in Schur Energy Expansion]
        Consequently, the exact leakage cross-covariance
        $\Sigma_{\lambda,Ik}$ has finite Cameron--Martin energy norm and is
        $O(\lambda^{-3})$ in that norm.  This is a condition on the deterministic
        response and covariance energy; it does not assert that infinite-dimensional
        Gaussian sample paths lie in $H_{Q_0}$ almost surely.
    \end{remark}

    % ============================================================
    %  §3  INVERSE-FREE SCHUR COMPLEMENTATION
    % ============================================================
    \section{Inverse-Free Schur Complementation in Hilbert Space}
    \label{sec:infdim}

    \subsection{Variational Shorting and Energy Bounds}
    \label{sec:infdim:variational}

    In finite dimensions, the Schur complement of the bulk covariance block $\Sigma_{II}$ is defined as $S_k \coloneqq \Sigma_{kk} - \Sigma_{kI} \Sigma_{II}^{-1} \Sigma_{Ik}$. Under stable and controllable dynamics, the bulk covariance $\Sigma_{II}$ is positive-definite and invertible, so the subtraction is well-defined. However, in infinite dimensions, the stationary covariance operator is trace-class and compact, meaning its restriction $\Sigma_{\lambda,II}$ lacks a bounded global inverse. The appropriate operator-theoretic replacement is the shorted operator: for a non-negative block operator, the Schur complement is recovered as the quadratic form of the short to the target subspace, with the inverse term replaced by a variational supremum \cite{krein1947selfadjoint,anderson1975shorted}. To lift the Schur complement to infinite dimensions in the present singular-conditioning problem, we formulate the projection residual directly using covariance-energy forms.

    Let $Q_0 \coloneqq \Sigma_{II}^{(0)}$ denote the fixed background bulk covariance under full decoupling ($\lambda = \infty$). We define the core Cameron--Martin space $H_{Q_0} \coloneqq \operatorname{Range}(Q_0^{1/2})$ and the core energy norm:
    \begin{equation}
        |u|_{Q_0} \coloneqq \|Q_0^{-1/2} u\| \quad \text{for all } u \in H_{Q_0}.
    \end{equation}
    By Assumption~\ref{assum:cm_stability}, the off-diagonal drift coupling vector satisfies $a_{Ik} \in H_{Q_0}$, and the restricted drift generator $\mathcal{A}_{II}$ generates an exponentially stable semigroup on $H_{Q_0}$.

    Under this assumption, the exact cross-covariance resolvent formula (derived in \Cref{app:block_cumulant_analysis}) satisfies:
    \begin{equation}
        \Sigma_{\lambda,Ik} = \frac{\varepsilon_0}{2\lambda^2} (\lambda I - \mathcal{A}_{II})^{-1} a_{Ik}.
    \end{equation}
    This resolvent response yields the Cameron--Martin energy bound on the cross-covariance block:
    \begin{equation}
        \label{eq:cross_norm_bound}
        |\Sigma_{\lambda,Ik}|_{Q_0} = O(\lambda^{-3}) \quad \text{as } \lambda \to \infty.
    \end{equation}

    To define the infinite-dimensional shorted residual, we introduce the variational Schur subtraction $R_\lambda$:
    \begin{equation}
        \label{eq:variational_schur_subtraction}
        R_\lambda \coloneqq \sup_{\langle u, \Sigma_{\lambda,II} u \rangle > 0} \frac{|\langle u, \Sigma_{\lambda,Ik} \rangle|^2}{\langle u, \Sigma_{\lambda,II} u \rangle}.
    \end{equation}
    Equivalently, if $\Sigma_\lambda$ is viewed as a non-negative block operator on $\mathbb{C}\phi_k\oplus\mathcal{H}_I$, then \eqref{eq:variational_schur_subtraction} is the one-dimensional Anderson--Trapp shorted-operator subtraction, namely
    \begin{equation}
        \label{eq:shorted_operator_scalar}
        \big\langle \phi_k,\mathcal{S}_{\mathbb{C}\phi_k}(\Sigma_\lambda)\phi_k\big\rangle
        =\Sigma_{\lambda,kk}-R_\lambda.
    \end{equation}
    The restriction to vectors with $\langle u,\Sigma_{\lambda,II}u\rangle>0$ simply removes null directions of the denominator; for a non-negative block operator, such directions do not change the supremum. Thus $R_\lambda$ is not an ambient inverse computation but the energy removed by shorting to the target coordinate \cite{anderson1975shorted,owhadi2015conditioning}.

    Since the target noise channel and bulk noise channel are independent, the bulk state decomposes into independent background and target-driven components, meaning the bulk covariance dominates the background covariance ($\Sigma_{\lambda,II} = Q_0 + \operatorname{Cov}(Y_I) \succeq Q_0$). Douglas range inclusion then guarantees that the variational subtraction is bounded:
    \begin{equation}
        0 \le R_\lambda \le |\Sigma_{\lambda,Ik}|_{Q_0}^2 = O(\lambda^{-6}).
    \end{equation}

    \subsection{Shorted Schur Residual Scaling}
    \label{sec:infdim:scaling}

    We now establish the main scaling theorem for the shorted Schur residual. The $L^2$ projection interpretation is secondary and follows from covariance geometry.

    \begin{theorem}[Hilbert Shorted Schur Invariant]
        \label{thm:inf_cancellation}
        Under Assumptions~\ref{assum:linear_second_moment} and \ref{assum:cm_stability}, the $L^2$-projection residual satisfies:
        \begin{equation}
            \label{eq:main_schur_sharp}
            S_{k,\lambda} \coloneqq \|X_k - P_{\mathcal{M}_I} X_k\|_{L^2}^2 = \frac{\varepsilon_0}{2\lambda^2} - R_\lambda,
        \end{equation}
        where the remainder satisfies $0 \le R_\lambda = O(\lambda^{-6})$. In particular, the shorted Schur invariant is:
        \begin{equation}
            \label{eq:m2_schur_value}
            m_2^{\mathrm{Schur}} = \lim_{\lambda\to\infty} \lambda^2 S_{k,\lambda} = \frac{\varepsilon_0}{2},
        \end{equation}
        and the reciprocal projection-residual scaling satisfies:
        \begin{equation}
            \label{eq:schur_scaling_inverse}
            S_{k,\lambda}^{-1} = \frac{2}{\varepsilon_0}\lambda^2 + O(\lambda^{-2}).
        \end{equation}
        This result depends only on covariance and orthogonal projection; it holds for any linear stationary model satisfying these assumptions, without Gaussianity.
    \end{theorem}
    \begin{proof}
        See \Cref{app:block_cumulant_analysis} for the full derivation.
    \end{proof}

    \subsection{Convergence of Finite-Dimensional Truncations}
    \label{sec:proofs:convergence}

    Let $\Pi_n: \mathcal{H} \to \mathcal{H}_n \coloneqq \operatorname{span}\{\phi_1, \ldots, \phi_n\}$ be the $n$-dimensional Galerkin projection. Write $\mathcal{A}_n^{(\lambda)} \coloneqq \Pi_n \mathcal{A}^{(\lambda)} \Pi_n$, $\mathcal{D}_n^{(\lambda)} \coloneqq \Pi_n \mathcal{D}^{(\lambda)} \Pi_n$, and let $\Sigma_{n,\lambda}$ solve the truncated Lyapunov equation $\mathcal{A}_n^{(\lambda)} \Sigma_{n,\lambda} + \Sigma_{n,\lambda} (\mathcal{A}_n^{(\lambda)})^* + \mathcal{D}_n^{(\lambda)} = 0$.

    \begin{proposition}[Galerkin Consistency of the Shorted Schur Invariant]
        \label{prop:galerkin_consistency}
        Under Assumption~\ref{assum:linear_second_moment}, let $\Pi_n:\mathcal{H}\to\mathcal{H}_n$ be finite-rank orthogonal Galerkin projections preserving the target-bulk block decomposition:
        \begin{equation}
            \Pi_n\phi_k=\phi_k,\qquad \Pi_n\mathcal{H}_I\subset\mathcal{H}_I,\qquad \Pi_n\to I\ \text{strongly on }\mathcal{H}.
        \end{equation}
        Assume the truncated regularized dynamics satisfy the uniform exponential stability and trace-class conditions stated in Proposition~\ref{prop:dynamic_galerkin} of \Cref{app:block_cumulant_analysis}.  Then the truncated covariance operators converge in trace norm:
        \begin{equation}
            \|\Sigma_{n,\lambda}-\Sigma_\lambda\|_{\mathcal{S}_1}\to0\quad\text{as }n\to\infty,
        \end{equation}
        and the truncated Schur invariant is exact at every level:
        \begin{equation}
            m_2^{\mathrm{Schur},(n)} \coloneqq \lim_{\lambda\to\infty} \lambda^2 S_{k,\lambda}^{(n)} = \frac{\varepsilon_0}{2} \quad \text{for all } n \ge k.
        \end{equation}
    \end{proposition}
    \begin{proof}
        See \Cref{app:block_cumulant_analysis} for the dynamic and spectral support.
    \end{proof}

    % ============================================================
    %  §4  FREDHOLM TRACE ANOMALIES AND GAUSSIAN FINITE PARTS
    % ============================================================
    \section{Fredholm Trace Anomalies and Gaussian Finite Parts}
    \label{sec:gaussian_zero}

    This section applies the inverse-free Schur construction to Gaussian
    Radon laws.  Disintegration supplies the exact-evidence measure; the new
    algebraic content is the cancellation between the Fredholm determinant
    term and the trace anomaly generated by the shorted covariance defect.

    \subsection{Radon Gaussian Application and Disintegration}
    \label{sec:gaussian:setup}

    \begin{assumption}[Gaussian Application and Affine Mean Channel]
        \label{assum:gaussian_clamp}
        In addition to Assumptions~\ref{assum:linear_second_moment} and \ref{assum:cm_stability}, let $\mu_\lambda = \mathcal{N}(m_\lambda, \Sigma_\lambda)$ be a Gaussian Radon law on $\mathcal{H}$ with injective trace-class covariance. Let $\mu_{\mathrm{obs}}$ be the canonical Gaussian disintegration member on the evidence hyperplane $X_k = x_k^*$. Suppose that a fixed deterministic forcing $b \in \mathcal{H}$ gives the stationary mean equation:
        \begin{equation}
            \mathcal{A}_{\cond} m_\lambda + b - \lambda \Pi_k (m_\lambda - x_k^* \phi_k) = 0, \quad \Pi_k \mathcal{A}_{\cond} = 0.
        \end{equation}
        We define the target deterministic offset constant:
        \begin{equation}
            c_\lambda \coloneqq \lambda(m_{\lambda,k} - x_k^*) = b_k,
        \end{equation}
        where $m_{\lambda,k} \coloneqq \langle m_\lambda, \phi_k \rangle$ and $b_k \coloneqq \langle b, \phi_k \rangle$.
    \end{assumption}

    Under the Gaussian Radon law, the coordinate projection $\pi_k: \mathcal{H} \to \mathbb{R}$ is continuous, so the target variable $X_k$ is a well-defined Borel random variable. The Polish structure of the Hilbert space guarantees the existence of a regular conditional probability distribution. 
    By selecting the weakly continuous version of the conditional law at the single point $t = x_k^*$ (as detailed in \Cref{app:gaussian_lift}), we obtain the unique pointwise disintegration member $\mu_{\mathrm{obs}}$. Under this measure, the target coordinate satisfies the exact target zeros:
    \begin{equation}
        \label{eq:exact_target_zeros}
        m_{\text{obs},k} - x_k^* = 0, \quad \Var_{\text{obs}}(X_k) = 0 \quad (\text{exact}).
    \end{equation}

    \subsection{Measurable Cameron--Martin Energy Projection}
    \label{sec:gaussian:predictor}

    Let $C_\lambda \coloneqq \Sigma_{\lambda,II}$ denote the Gaussian bulk covariance operator on $\mathcal{H}_I$. Because $C_\lambda$ is trace-class and compact, it lacks a bounded global inverse. We define the bulk Cameron--Martin space $H_{C_\lambda} \coloneqq \operatorname{Range}(C_\lambda^{1/2})$ and the corresponding energy norm.

    The leakage cross-covariance vector satisfies $\Sigma_{\lambda,Ik} \in H_{C_\lambda}$ with energy norm $\|\Sigma_{\lambda,Ik}\|_{H_{C_\lambda}} = O(\lambda^{-3})$. This range inclusion allows us to define the energy projection $\mu_{k|I}(X_I) \coloneqq \E[X_k \mid X_I]$ as a measurable Wiener functional:
    \begin{equation}
        \mu_{k|I}(X_I) = m_{\lambda,k} + W_{\Sigma_{\lambda,Ik}}(X_I - m_{\lambda,I}),
    \end{equation}
    which is the unique linear $L^2$ predictor. The estimates in \Cref{app:gaussian_lift} imply that, for all sufficiently large $\lambda$, the bulk marginals $\mu_{\text{obs},I}$ and $\mu_{\lambda,I}$ are equivalent and have finite bulk relative entropy:
    \begin{equation}
        \KL(\mu_{\text{obs},I} \parallel \mu_{\lambda,I}) = O(\lambda^{-4}) \to 0 \quad \text{as } \lambda \to \infty.
    \end{equation}

    \subsection{Gaussian Fredholm Finite-Part Functional}
    \label{sec:gaussian:functional}

    Although the joint KL divergence diverges under exact conditioning, the
    block-operator decomposition isolates a finite Fredholm part.

    \begin{definition}[Gaussian Fredholm Finite-Part Functional]
        \label{def:unified_action}
        Whenever $S_{k,\lambda}>0$ and $\mu_{\text{obs},I}$ is equivalent to $\mu_{\lambda,I}$, we define the Gaussian Fredholm finite-part functional of the conditioned law:
        \begin{equation}
            \label{eq:unified_action}
            \begin{aligned}
                \mathfrak{J}_\lambda(\mu_{\mathrm{obs}}; \mu_\lambda) \coloneqq &\KL(\mu_{\mathrm{obs},I} \parallel \mu_{\lambda,I}) \\
                &+ \frac{1}{2} S_{k,\lambda}^{-1} \E_{\mu_{\mathrm{obs}}}\left[ \left( X_k - \mu_{k|I}(X_I) \right)^2 \right].
            \end{aligned}
        \end{equation}
        This is the finite part of the joint KL divergence after subtracting the Dirac normalization singularity; the remaining covariance contribution is controlled by Fredholm determinant and trace terms.
    \end{definition}

    \paragraph{Weak scheme-independence.}
    For an evidence level $t\in\mathbb{R}$, write $\mu_\lambda^t$ for the weakly continuous conditional law from \Cref{app:gaussian_lift} and set
    \[
    \mathfrak{J}_\lambda(t)\coloneqq
    \mathfrak{J}_\lambda(\mu_\lambda^t;\mu_\lambda).
    \]
    Then, for any two evidence levels $t_1,t_2$ for which the finite part is defined,
    \begin{equation}
        \label{eq:weak_scheme_independence}
        \mathfrak{J}_\lambda(t_2)-\mathfrak{J}_\lambda(t_1)
        =
        \frac{(t_2-m_{\lambda,k})^2-(t_1-m_{\lambda,k})^2}
        {2\Sigma_{\lambda,kk}}.
    \end{equation}
    In the calibrated scaling $\Sigma_{\lambda,kk}=\varepsilon_0/(2\lambda^2)$, if
    $b_i=\lambda(m_{\lambda,k}-t_i)$, this becomes
    \[
    \mathfrak{J}_\lambda(t_2)-\mathfrak{J}_\lambda(t_1)
    =
    \frac{b_2^2-b_1^2}{\varepsilon_0}.
    \]
    \begin{proof}
        Let $\Delta_t\coloneqq t-m_{\lambda,k}$ and
        \[
        \rho_\lambda
        \coloneqq
        \frac{\|\Sigma_{\lambda,Ik}\|_{H_{C_\lambda}}^2}{\Sigma_{\lambda,kk}}
        =
        1-\frac{S_{k,\lambda}}{\Sigma_{\lambda,kk}}.
        \]
        The Gaussian entropy computation in \Cref{app:gaussian_lift} gives
        \[
        \KL(\mu_{\lambda,I}^t\parallel\mu_{\lambda,I})
        =
        \frac{1}{2}\frac{\Delta_t^2}{\Sigma_{\lambda,kk}}\rho_\lambda
        -\frac{1}{2}\log(1-\rho_\lambda)-\frac{1}{2}\rho_\lambda .
        \]
        The conditional-residual term in \eqref{eq:unified_action} is
        \[
        \frac{1}{2S_{k,\lambda}}
        \E_{\mu_\lambda^t}\!\left[
        (X_k-\mu_{k|I}(X_I))^2
        \right]
        =
        \frac{1}{2}\rho_\lambda
        +
        \frac{1}{2}\frac{\Delta_t^2}{\Sigma_{\lambda,kk}}(1-\rho_\lambda).
        \]
        Adding these two identities cancels the $\rho_\lambda/2$ terms and yields
        \[
        \mathfrak{J}_\lambda(t)
        =
        \frac{(t-m_{\lambda,k})^2}{2\Sigma_{\lambda,kk}}
        -\frac{1}{2}\log(1-\rho_\lambda).
        \]
        The logarithmic term depends on the reference covariance and the conditioning
        scale, but not on the evidence value $t$. Hence it cancels in the difference
        between $t_2$ and $t_1$, proving \eqref{eq:weak_scheme_independence}. The same
        argument shows weak independence from the mollifying kernel: replacing the
        Gaussian Dirac mollifier in \Cref{prop:mollification_limit} by any centered
        approximate identity with finite second moment changes the extracted finite
        part only by an additive term depending on the kernel, $\lambda$, and
        $S_{k,\lambda}$, not on $t$. Therefore all evidence differences
        $\Delta\mathfrak{J}_\lambda$ are kernel-independent.
    \end{proof}

    \subsection{Gaussian Main Theorem}
    \label{sec:gaussian:theorem}

    We now establish the main Fredholm regularization and cancellation theorems for the finite-part functional.

    \begin{theorem}[Gaussian Fredholm Finite-Part Regularization]
        \label{thm:finite_part_regularization}
        Under Assumption~\ref{assum:gaussian_clamp}, there exists $\lambda_0>0$ such that, for $\lambda\ge\lambda_0$, the Gaussian Fredholm finite-part functional under the conditioned law $\mu_{\mathrm{obs}}$ is well-defined and satisfies:
        \begin{equation}
            \label{eq:unified_regularization}
            \mathfrak{J}_\lambda(\mu_{\mathrm{obs}}; \mu_\lambda) = \frac{b_k^2}{\varepsilon_0} + O(\lambda^{-4}).
        \end{equation}
        In particular, the functional remains bounded at $O(1)$ and converges to:
        \begin{equation}
            \lim_{\lambda\to\infty} \mathfrak{J}_\lambda(\mu_{\mathrm{obs}}; \mu_\lambda) = \frac{b_k^2}{\varepsilon_0}.
        \end{equation}
    \end{theorem}
    \begin{proof}
        See \Cref{app:gaussian_lift} for the full derivation.
    \end{proof}

    \begin{corollary}[Exact Decoupled-Centered Cancellation]
        \label{cor:exact_pole_zero}
        Under Assumption~\ref{assum:gaussian_clamp}, if the target coordinate is dynamically decoupled from the bulk ($a_{Ik} = 0$) and the evidence is centered at the stationary mean ($x_k^* = m_{\lambda,k}$), then:
        \begin{equation}
            \mathfrak{J}_\lambda(\mu_{\mathrm{obs}}; \mu_\lambda) = 0 \quad \text{for every } \lambda > 0.
        \end{equation}
    \end{corollary}
    \begin{proof}
        The decoupling implies $\Sigma_{\lambda,Ik} = 0$, so $\KL(\mu_{\text{obs},I} \parallel \mu_{\lambda,I}) = 0$. The centering makes the conditional residual zero under the conditioning, so both terms in \eqref{eq:unified_action} vanish.
    \end{proof}

    % ============================================================
    %  §5  ELLIPTIC GAUSSIAN FIELD EXAMPLE
    % ============================================================
    \section{Elliptic Gaussian Field Example}
    \label{sec:spde}

    We now record a standard elliptic Gaussian field example to show that the abstract Hilbert-space assumptions are nonempty and natural for covariance operators arising from stochastic evolution equations and elliptic Gaussian priors \cite{daprato2014stochastic,lord2014introduction,stuart2010inverse}.

    \subsection{Elliptic Gaussian field on \texorpdfstring{$[0,1]$}{[0,1]}}

    Let
    \[
    L=-\frac{d^2}{dx^2}+\kappa^2
    \]
    on $L^2(0,1)$ with Dirichlet boundary conditions and $\kappa>0$. Then $\mathcal{A}_0=-L$ with domain $H^2(0,1)\cap H_0^1(0,1)$ is dissipative, $L$ is positive self-adjoint and sectorial, and $e^{-tL}$ is an exponentially stable analytic contraction semigroup. The inverse covariance
    \[
    Q_0=L^{-1}
    \]
    is compact, positive, self-adjoint, and trace class. With eigenfunctions $e_j(x)=\sqrt{2}\sin(\pi jx)$, its eigenvalues are
    \[
    q_j=(\pi^2j^2+\kappa^2)^{-1}.
    \]
    Thus $Q_0$ is a typical infinite-dimensional covariance operator: it is compact and trace class, has no bounded inverse on the ambient space, and carries its inverse only as a Cameron--Martin energy form \cite{bogachev1998gaussian,simon2005trace}.

    \subsection{Cameron--Martin range and covariance energy}

    The Cameron--Martin space associated with $Q_0$ is
    \[
    H_{Q_0}=\operatorname{Range}(Q_0^{1/2})=H_0^1(0,1),
    \]
    with equivalent energy norm
    \[
    \|h\|_{Q_0}^2
    =\langle h,Lh\rangle_{L^2}
    =\int_0^1 \bigl(|h'(x)|^2+\kappa^2|h(x)|^2\bigr)\,dx.
    \]
    This identifies the variational Schur projection in a Sobolev-energy form, as in the standard covariance-energy representation of Gaussian Hilbert priors \cite{bogachev1998gaussian,stuart2010inverse}. The covariance inverse does not act as a bounded operator on $L^2(0,1)$; it acts only through the energy form on its Cameron--Martin domain.

    \subsection{Spectral sensor conditioning}

    Fix a target eigenmode $e_k$ and define the observed coordinate
    \[
    X_k=\langle X,e_k\rangle_{L^2}.
    \]
    The unperturbed target variance is
    \[
    \Sigma_{0,kk}=q_k=(\pi^2k^2+\kappa^2)^{-1}.
    \]
    A regularized restoring drift on this coordinate produces the exact normalization
    \[
    \Sigma_{\lambda,kk}=\frac{\varepsilon_0}{2\lambda^2}-R_\lambda,
    \qquad R_\lambda=O(\lambda^{-6}),
    \]
    under the block assumptions of \Cref{sec:infdim}. Off-diagonal bounded perturbations of the drift operator generate the cross-covariance vector entering the variational Schur residual. The resulting subtraction is evaluated by the Cameron--Martin energy form above, not by a global inverse on $L^2(0,1)$.

    This example shows that the abstract theorem applies to a genuine Gaussian field. The residual is controlled by a Sobolev energy on $H_0^1(0,1)$ while the ambient covariance remains compact.

    % ============================================================
    %  \S6  FINITE-RANK SHORTING AND EXACT FREDHOLM TRACE CANCELLATION
    % ============================================================
    \section{Finite-Rank Shorting and Exact Fredholm Trace Cancellation}
    \label{sec:finite_rank}
    \label{sec:discussion}

We have resolved the rank-one block-operator obstruction: a compact
free covariance block need not be inverted, because the Schur subtraction is
the Anderson--Trapp shorted residual on the Cameron--Martin energy range.
Finite-rank targets require the same idea in operator-valued form, plus an
exact matrix/Fredholm identity for the Gaussian finite part.  Here Galerkin
approximation remains purely variational: truncation restricts the
Anderson--Trapp supremum to nested finite-dimensional test spaces.  The
limitation is not finite rank, but infinite-rank observation, where Fredholm
determinants, trace-class perturbation hypotheses, and operator-valued finite
parts need separate treatment.

\subsection{Matrix algebraic extension}
\label{sec:finite_rank:matrix_extension}

For an $m$-dimensional observed subspace $E=\operatorname{span}\{\phi_1,\ldots,\phi_m\}$, the scalar target block $\Sigma_{\lambda,kk}$ is replaced by the covariance matrix $\Sigma_{\lambda,EE}\in\mathbb{R}^{m\times m}$. The scalar calibration becomes a matrix Lyapunov balancing problem.

\begin{theorem}[Matrix Lyapunov Calibration]
  \label{thm:matrix_lyapunov}
  Let $E$ be an $m$-dimensional target subspace. Suppose the regularized drift and diffusion on $E$ are given by $A_{\lambda,E} = -\lambda \Lambda_E$ and $D_{\lambda,E} = \frac{1}{\lambda} \Delta_E$, where $\Lambda_E, \Delta_E \in \mathbb{R}^{m \times m}$ are strictly positive definite matrices. Let $\Sigma_{\lambda,EE}$ be the unique positive definite solution to the matrix Lyapunov equation:
  \begin{equation}
    \label{eq:discussion_matrix_lyapunov}
    A_{\lambda,E}\Sigma_{\lambda,EE}
    +\Sigma_{\lambda,EE}A_{\lambda,E}^{*}
    +D_{\lambda,E}=0.
  \end{equation}
  Then $\Sigma_{\lambda,EE}$ scales exactly as $\lambda^{-2}$. Specifically, the normalized matrix $\widetilde{\Sigma}_{EE} \coloneqq \lambda^2 \Sigma_{\lambda,EE}$ is independent of $\lambda$ and is the unique positive definite solution to the scale-free Lyapunov equation:
  \begin{equation}
    \label{eq:scale_free_lyapunov}
    \Lambda_E \widetilde{\Sigma}_{EE} + \widetilde{\Sigma}_{EE} \Lambda_E^* = \Delta_E.
  \end{equation}
\end{theorem}
\begin{proof}
  Substituting $A_{\lambda,E} = -\lambda \Lambda_E$ and $D_{\lambda,E} = \frac{1}{\lambda} \Delta_E$ into \eqref{eq:discussion_matrix_lyapunov} yields:
  \[
    -\lambda \Lambda_E \Sigma_{\lambda,EE} - \lambda \Sigma_{\lambda,EE} \Lambda_E^* + \frac{1}{\lambda} \Delta_E = 0.
  \]
  Multiplying the entire equation by $\lambda$ gives:
  \[
    \Lambda_E (\lambda^2 \Sigma_{\lambda,EE}) + (\lambda^2 \Sigma_{\lambda,EE}) \Lambda_E^* = \Delta_E.
  \]
  Defining $\widetilde{\Sigma}_{EE} \coloneqq \lambda^2 \Sigma_{\lambda,EE}$, we recover \eqref{eq:scale_free_lyapunov}. Since $\Lambda_E$ is positive definite, its spectrum lies in the open right half-plane, making $-\Lambda_E$ Hurwitz. Because $\Delta_E$ is positive definite, standard continuous Lyapunov theory guarantees that \eqref{eq:scale_free_lyapunov} admits a unique positive definite solution $\widetilde{\Sigma}_{EE}$. Therefore, the exact scaling $\Sigma_{\lambda,EE} = \lambda^{-2} \widetilde{\Sigma}_{EE}$ holds for all $\lambda > 0$, properly generalizing the scalar calibration condition $\Sigma_{\lambda,kk} = \varepsilon_0 / (2\lambda^2)$.
\end{proof}

\subsection{Inverse-Free Operator Residuals and Variational Loewner Monotonicity}
\label{sec:discussion:operator-valued-residuals}

For an $m$-dimensional target space $E$, the Schur residual is a
positive-semidefinite operator on $E$.  Since the free-block covariance
$C_\lambda\coloneqq\Sigma_{\lambda,II}$ is compact and trace class on the
infinite-dimensional free space, it has no bounded global inverse.  Thus
the formal expression
\[
  \Sigma_{\lambda,EI}C_\lambda^{-1}\Sigma_{\lambda,IE}
\]
is not used as a definition.  The residual is defined instead by the
Anderson--Trapp variational shorting formula.

\begin{theorem}[Inverse-free operator residual and Galerkin Loewner monotonicity]
\label{thm:operator_residual}
Let $\mathcal H=E\oplus\mathcal H_I$, where $E$ is finite dimensional.
Let $C_\lambda=\Sigma_{\lambda,II}$ be a non-negative trace-class
covariance operator on $\mathcal H_I$, and set
\[
  B_\lambda\coloneqq \Sigma_{\lambda,IE}:E\to\mathcal H_I .
\]
Assume the Cameron--Martin range condition
\[
  B_\lambda E\subset H_{C_\lambda}
  \coloneqq \operatorname{Ran}(C_\lambda^{1/2}).
\]
Define the shorted residual $R_{\lambda,E}$ on $E$ by
\begin{equation}
\label{eq:discussion_operator_residual_form}
  \langle v,R_{\lambda,E}v\rangle_E
  =
  \sup_{\substack{u\in\mathcal H_I\\ \langle u,C_\lambda u\rangle>0}}
  \frac{|\langle u,B_\lambda v\rangle_{\mathcal H_I}|^2}
       {\langle u,C_\lambda u\rangle_{\mathcal H_I}},
  \qquad v\in E .
\end{equation}
Then $R_{\lambda,E}$ is a well-defined bounded positive-semidefinite
operator on $E$.  If
\[
  \|B_\lambda v\|_{H_{C_\lambda}}
  \le M\lambda^{-3}\|v\|_E,
  \qquad v\in E,
\]
then
\[
  \|R_{\lambda,E}\|_{\mathcal L(E)}=O(\lambda^{-6}).
\]

Let $(\mathcal H_n)_{n\ge1}$ be an increasing sequence of
finite-dimensional subspaces of $\mathcal H_I$ with dense union.  Define
the Galerkin residuals by the restricted variational problems
\begin{equation}
\label{eq:galerkin_variational_residual}
  \langle v,R_{\lambda,E}^{(n)}v\rangle_E
  \coloneqq
  \sup_{\substack{u\in\mathcal H_n\\ \langle u,C_\lambda u\rangle>0}}
  \frac{|\langle u,B_\lambda v\rangle_{\mathcal H_I}|^2}
       {\langle u,C_\lambda u\rangle_{\mathcal H_I}},
  \qquad v\in E .
\end{equation}
Then
\[
  0\preceq R_{\lambda,E}^{(n)}
  \preceq R_{\lambda,E}^{(n+1)}
  \preceq R_{\lambda,E}
\]
in the Loewner order, and
\[
  R_{\lambda,E}^{(n)}
  \longrightarrow
  R_{\lambda,E}
\]
in operator norm on $E$.
\end{theorem}

\begin{proof}
Fix $v\in E$ and write $y_v=B_\lambda v$.  By the range assumption,
$y_v\in H_{C_\lambda}$.  Thus the Cameron--Martin norm
$\|y_v\|_{H_{C_\lambda}}$ is finite.  Equivalently, the quadratic-form
Cauchy--Schwarz inequality gives
\[
  |\langle u,y_v\rangle_{\mathcal H_I}|^2
  \le
  \|y_v\|_{H_{C_\lambda}}^2
  \langle u,C_\lambda u\rangle_{\mathcal H_I},
  \qquad u\in\mathcal H_I .
\]
Hence the supremum in
\eqref{eq:discussion_operator_residual_form} is finite and equals
\[
  \|y_v\|_{H_{C_\lambda}}^2 .
\]
Therefore $v\mapsto \langle v,R_{\lambda,E}v\rangle_E$ is a finite
non-negative quadratic form on the finite-dimensional space $E$, so it is
represented by a unique positive-semidefinite operator $R_{\lambda,E}$.
The same argument applies to each restricted supremum and defines
$R_{\lambda,E}^{(n)}$.

Since
\[
  \mathcal H_n\subset \mathcal H_{n+1}\subset\mathcal H_I,
\]
the admissible class in the restricted variational problem increases
with $n$.  Hence, for every $v\in E$,
\[
  \langle v,R_{\lambda,E}^{(n)}v\rangle_E
  \le
  \langle v,R_{\lambda,E}^{(n+1)}v\rangle_E
  \le
  \langle v,R_{\lambda,E}v\rangle_E .
\]
This is precisely the Loewner monotonicity.

It remains to identify the limit.  Let
\[
  q_n(v)=\langle v,R_{\lambda,E}^{(n)}v\rangle_E,
  \qquad
  q(v)=\langle v,R_{\lambda,E}v\rangle_E .
\]
The sequence $q_n(v)$ is increasing and bounded above by $q(v)$.
Conversely, fix any admissible $u\in\mathcal H_I$ with
$\langle u,C_\lambda u\rangle>0$.  Since $\bigcup_n\mathcal H_n$ is
dense, there exist $u_n\in\mathcal H_n$ with $u_n\to u$ in
$\mathcal H_I$.  Because $C_\lambda$ is bounded and $y_v$ is fixed,
\[
  \langle u_n,C_\lambda u_n\rangle
  \to
  \langle u,C_\lambda u\rangle,
  \qquad
  \langle u_n,y_v\rangle
  \to
  \langle u,y_v\rangle .
\]
Thus every value of the full Rayleigh quotient can be approximated by
values taken over Galerkin subspaces.  Taking suprema gives
\[
  \lim_{n\to\infty}q_n(v)=q(v),
  \qquad v\in E .
\]
Since $E$ is finite dimensional, pointwise convergence of the quadratic
forms implies convergence of their representing matrices in operator
norm.

Finally, the assumed Cameron--Martin energy estimate gives
\[
  \langle v,R_{\lambda,E}v\rangle_E
  =
  \|B_\lambda v\|_{H_{C_\lambda}}^2
  \le
  M^2\lambda^{-6}\|v\|_E^2,
\]
and hence
\[
  \|R_{\lambda,E}\|_{\mathcal L(E)}
  \le M^2\lambda^{-6}.
\]
\end{proof}

\begin{remark}[Why strict Loewner growth is not expected]
\label{rem:strict_monotonicity}
The monotonicity in \Cref{thm:operator_residual} is non-decreasing
Loewner monotonicity.  Strict Loewner growth is not structurally
guaranteed.

Indeed, suppose that the Galerkin space is enlarged by a finite-dimensional
increment
\[
  \Delta\mathcal H_n
  =
  \mathcal H_{n+1}\ominus\mathcal H_n
\]
of dimension $k$.  In the Cameron--Martin energy geometry, the additional
Rayleigh--Ritz information available at the next step has rank at most
$k$.  Consequently
\[
  \operatorname{rank}
  \bigl(R_{\lambda,E}^{(n+1)}-R_{\lambda,E}^{(n)}\bigr)
  \le k .
\]
If the Galerkin scheme adds only one new direction at each step, then the
increment is at most rank one.  For a target space of dimension $m>1$, a
rank-one positive-semidefinite increment cannot be positive definite on
$E$.  Therefore
\[
  R_{\lambda,E}^{(n+1)}-R_{\lambda,E}^{(n)}\succ0
\]
is impossible for one-dimensional Galerkin updates when $m>1$.

Strict positivity of an increment requires both enough new Galerkin
directions, $k\ge m$, and full-rank visibility of the unresolved
cross-covariance defect across all target directions.
\end{remark}

\subsection{General observation operators and misalignment}
\label{sec:discussion:misalignment}

The scalar proof in this paper uses a block decomposition aligned with a single observed coordinate. However, typical finite-rank observation operators $L:\mathcal H\to\mathbb R^m$ select a subspace $E$ that does not commute with the free prior covariance, meaning the target-free split is not an eigenbasis split. This geometric misalignment generates cross-coupling that requires a precise resolvent analysis.

\begin{theorem}[Misaligned Covariance Coupling]
  \label{thm:misaligned_coupling}
  Let $P_E$ be an orthogonal projection onto an $m$-dimensional observed subspace $E$, and let $\mathcal{H}_I = E^\perp$. Let the regularized dynamics be decomposed such that the target block matches \Cref{thm:matrix_lyapunov} with $A_{\lambda,E} = -\lambda \Lambda_E$, the cross-noise is zero ($D_{\lambda,IE} = 0$), and the off-diagonal drift coupling is denoted by $A_{\lambda,IE}$.
  Assume that, for some $\lambda_0>0$,
  \[
    \sup_{\lambda\ge\lambda_0}
    \|A_{\lambda,IE}\|_{\mathcal L(E,H_{Q_0})}<\infty.
  \]
  Assume also that the restricted drift $A_{II}$ generates an exponentially stable semigroup on $H_{Q_0}$ and that there exist $K,\gamma>0$ such that
  \[
    \|e^{-t\Lambda_E^*}\|_{\mathcal L(E)}
    \le
    K e^{-\gamma t},
    \qquad t\ge0.
  \]
  Then the cross-covariance satisfies the exact identity:
  \begin{equation}
    \label{eq:misaligned_resolvent_identity}
    \Sigma_{\lambda,IE} = \int_0^\infty e^{t A_{II}} A_{\lambda,IE} \Sigma_{\lambda,EE} e^{t A_{\lambda,E}^*} dt.
  \end{equation}
  Furthermore, the misaligned cross-covariance energy scales as $\|\Sigma_{\lambda,IE} v\|_{H_{Q_0}} = O(\lambda^{-3})$ for any $v \in E$.
\end{theorem}
\begin{proof}
  The off-diagonal block of the global Lyapunov equation $\mathcal{A}^{(\lambda)} \Sigma_\lambda + \Sigma_\lambda (\mathcal{A}^{(\lambda)})^* + \mathcal{D}^{(\lambda)} = 0$ yields the Sylvester equation:
  \[
    A_{II} \Sigma_{\lambda,IE} + \Sigma_{\lambda,IE} A_{\lambda,E}^* + A_{\lambda,IE} \Sigma_{\lambda,EE} = 0.
  \]
  Since $A_{II}$ generates an exponentially stable semigroup $e^{t A_{II}}$ and the assumed bound controls $e^{tA_{\lambda,E}^*}=e^{-\lambda t\Lambda_E^*}$, the unique solution is given by the integral representation \eqref{eq:misaligned_resolvent_identity}.

  To bound the Cameron--Martin norm, fix $v \in E$. We have $e^{t A_{\lambda,E}^*} v = e^{-\lambda t \Lambda_E^*} v$. By \Cref{thm:matrix_lyapunov}, $\Sigma_{\lambda,EE} = O(\lambda^{-2})$. By the uniform coupling assumption, choose $C$ such that $\|A_{\lambda,IE} w\|_{H_{Q_0}} \le C \|w\|_E$ for all $w \in E$ and all $\lambda\ge\lambda_0$. Using the uniform exponential stability of $e^{t A_{II}}$ on $H_{Q_0}$, namely $\|e^{t A_{II}}\|_{\mathcal{L}(H_{Q_0})} \le M e^{-\omega t}$ for some $M \ge 1, \omega > 0$, we estimate:
  \begin{align*}
    \|\Sigma_{\lambda,IE} v\|_{H_{Q_0}}
    &\le \int_0^\infty \|e^{t A_{II}}\|_{\mathcal{L}(H_{Q_0})} \|A_{\lambda,IE} \Sigma_{\lambda,EE} e^{-\lambda t \Lambda_E^*} v\|_{H_{Q_0}} dt \\
    &\le \int_0^\infty M e^{-\omega t} C \|\Sigma_{\lambda,EE}\|_{\mathcal{L}(E)} \|e^{-\lambda t \Lambda_E^*}\|_{\mathcal{L}(E)} \|v\|_E dt.
  \end{align*}
  Using $\|e^{-\lambda t\Lambda_E^*}\|_{\mathcal{L}(E)} \le K e^{-\lambda\gamma t}$, we obtain:
  \begin{align*}
    \|\Sigma_{\lambda,IE} v\|_{H_{Q_0}}
    &\le M C K \|\Sigma_{\lambda,EE}\|_{\mathcal{L}(E)} \|v\|_E \int_0^\infty e^{-(\omega + \lambda \gamma) t} dt \\
    &= M C K \|\Sigma_{\lambda,EE}\|_{\mathcal{L}(E)} \|v\|_E \frac{1}{\omega + \lambda \gamma}.
  \end{align*}
  Since $\|\Sigma_{\lambda,EE}\|_{\mathcal{L}(E)} = O(\lambda^{-2})$ and $(\omega + \lambda \gamma)^{-1} = O(\lambda^{-1})$, we conclude that $\|\Sigma_{\lambda,IE} v\|_{H_{Q_0}} = O(\lambda^{-3})$. Combined with \Cref{thm:operator_residual}, this verifies that the shorted-operator residual maintains the $O(\lambda^{-6})$ scaling even in the presence of observation misalignment.
\end{proof}

\subsection{Multidimensional Fredholm Trace-Anomaly Cancellation}
\label{sec:discussion:multidim_functional}

With the finite-rank covariance bounds in
\Cref{thm:matrix_lyapunov,thm:operator_residual,thm:misaligned_coupling} in
place, the rank-one finite-part functional from \Cref{sec:gaussian_zero}
becomes a matrix problem.  The determinant contribution is no longer a scalar
logarithm expansion; it is a Fredholm determinant reduced to an
$m\times m$ matrix by the Weinstein--Aronszajn identity.  The trace anomaly
then cancels exactly against the matrix-valued shorted residual term.

The scalar renormalized functional can be fully generalized to an $m$-dimensional observation subspace $E$. Let the exact evidence be $X_E = x_E^* \in \mathbb{R}^m$. The true Schur residual matrix is $S_{\lambda,E} \coloneqq \Sigma_{\lambda,EE} - R_{\lambda,E} \in \mathbb{R}^{m \times m}$, where $R_{\lambda,E}$ is the shorted-operator residual from \Cref{thm:operator_residual}. Since $R_{\lambda,E}^{(n)}\nearrow R_{\lambda,E}$ in the Loewner order, the Galerkin residual matrices
\[
S_{\lambda,E}^{(n)}
=
\Sigma_{\lambda,EE}-R_{\lambda,E}^{(n)}
\]
decrease in the Loewner order to
\[
S_{\lambda,E}
=
\Sigma_{\lambda,EE}-R_{\lambda,E}.
\]
They are not claimed to decrease strictly.  Positivity of the limiting
matrix is guaranteed by the condition
\[
\rho_{\lambda,E}
=
\Sigma_{\lambda,EE}^{-1/2}
R_{\lambda,E}
\Sigma_{\lambda,EE}^{-1/2}
<I.
\]

\begin{definition}[Multidimensional Fredholm Finite-Part Functional]
  \label{def:multidim_unified_action}
  Whenever $S_{\lambda,E} \succ 0$ and $\mu_{\text{obs},I}$ is equivalent to $\mu_{\lambda,I}$, the multidimensional Gaussian Fredholm finite-part functional is defined by evaluating the conditional residual in the metric of the shorted Schur matrix:
  \begin{equation}
    \label{eq:multidim_unified_action}
    \begin{aligned}
      \mathfrak{J}_{\lambda,E}(\mu_{\mathrm{obs}}; \mu_\lambda) \coloneqq &\KL(\mu_{\mathrm{obs},I} \parallel \mu_{\lambda,I}) \\
      &+ \frac{1}{2} \E_{\mu_{\mathrm{obs}}}\left[ \left( X_E - \mu_{E|I}(X_I) \right)^T S_{\lambda,E}^{-1} \left( X_E - \mu_{E|I}(X_I) \right) \right],
    \end{aligned}
  \end{equation}
  where $\mu_{E|I}(X_I) \coloneqq m_{\lambda,E} + W_{\Sigma_{\lambda,IE}}(X_I - m_{\lambda,I})$ is the measurable Cameron--Martin energy projection on $\mathbb{R}^m$, with the Wiener integral evaluated componentwise.
\end{definition}

To establish the multidimensional limit, we define the dimensionless coupling matrix $\rho_{\lambda,E} \coloneqq \Sigma_{\lambda,EE}^{-1/2} R_{\lambda,E} \Sigma_{\lambda,EE}^{-1/2} \in \mathbb{R}^{m \times m}$. By the energy bound in \Cref{thm:operator_residual}, $\|\rho_{\lambda,E}\| = O(\lambda^{-4})$.

\begin{theorem}[Exact Fredholm Trace and Matrix Determinant Cancellation]
  \label{thm:multidim_cancellation}
  For any evidence level $t \in \mathbb{R}^m$, let $\mu_{\lambda}^t$ be the weakly continuous conditional law on $X_E = t$. The infinite-dimensional KL divergence and the matrix-valued shorted residual expectation undergo exact cancellation of all cross-coupling traces and quadratic forms, yielding:
  \begin{equation}
    \label{eq:multidim_exact_cancellation}
    \mathfrak{J}_{\lambda,E}(\mu_\lambda^t; \mu_\lambda) = \frac{1}{2} (t - m_{\lambda,E})^T \Sigma_{\lambda,EE}^{-1} (t - m_{\lambda,E}) - \frac{1}{2}\log\det(I-\rho_{\lambda,E}).
  \end{equation}
\end{theorem}
\begin{proof}
  Let $C_\lambda \coloneqq \Sigma_{\lambda,II}$. Under $\mu_\lambda^t$, the free coordinates follow $\mathcal{N}(\mu_{I|E}(t), \Sigma_{I|E})$. By the Feldman--H\'{a}jek formula for trace-class perturbations, evaluating the KL divergence relies on the operator $K \coloneqq C_\lambda^{-1/2} \Sigma_{\lambda,IE} \Sigma_{\lambda,EE}^{-1/2}$. By the Weinstein--Aronszajn identity, the non-zero spectrum of the infinite-dimensional operator $KK^*$ matches that of the $m \times m$ matrix $K^*K = \rho_{\lambda,E}$. This reduces the infinite-dimensional Fredholm determinant and trace exactly to the matrix manifold:
  \[
    \KL(\mu_{\lambda,I}^t \parallel \mu_{\lambda,I}) = \frac{1}{2} \left[ (t - m_{\lambda,E})^T \Sigma_{\lambda,EE}^{-1} R_{\lambda,E} \Sigma_{\lambda,EE}^{-1} (t - m_{\lambda,E}) - \log\det(I - \rho_{\lambda,E}) - \tr(\rho_{\lambda,E}) \right].
  \]
  For the residual expectation term, let $Z \coloneqq t - \mu_{E|I}(X_I)$. Using block matrix identities, the conditional mean is $\E_{\mu_\lambda^t}[Z] = S_{\lambda,E} \Sigma_{\lambda,EE}^{-1} (t - m_{\lambda,E})$ and the conditional variance is $\Var_{\mu_\lambda^t}(Z) = R_{\lambda,E} \Sigma_{\lambda,EE}^{-1} S_{\lambda,E}$. Expanding the quadratic form $\frac{1}{2}\E[Z^T S_{\lambda,E}^{-1} Z]$ gives:
  \[
    \frac{1}{2} \left[ (t - m_{\lambda,E})^T \Sigma_{\lambda,EE}^{-1} (t - m_{\lambda,E}) - (t - m_{\lambda,E})^T \Sigma_{\lambda,EE}^{-1} R_{\lambda,E} \Sigma_{\lambda,EE}^{-1} (t - m_{\lambda,E}) + \tr(\rho_{\lambda,E}) \right].
  \]
  Summing the KL divergence and the quadratic residual form exactly cancels both the $\pm \frac{1}{2}\tr(\rho_{\lambda,E})$ trace anomalies and the cross-covariance residual matrices, isolating \eqref{eq:multidim_exact_cancellation}.
\end{proof}

\begin{theorem}[Multidimensional Fredholm Finite-Part Regularization]
  \label{thm:multidim_regularization}
  Assume the $m$-dimensional regularized drift on $E$ is centered at the observed value:
  \[
    A_{\lambda,E}(m_{\lambda,E}-x_E^*)+b_E=0,
    \qquad
    A_{\lambda,E}=-\lambda\Lambda_E.
  \]
  Defining the normalized displacement $\delta_E \coloneqq \Lambda_E^{-1} b_E \in \mathbb{R}^m$, the mean deterministic offset satisfies
  \[
    m_{\lambda,E}-x_E^*
    =
    \lambda^{-1}\Lambda_E^{-1}b_E
    =
    \lambda^{-1}\delta_E.
  \]

  As $\lambda \to \infty$, the multidimensional Fredholm finite-part functional converges to a well-defined bounded limit governed by the scale-free Lyapunov solution $\widetilde{\Sigma}_{EE}$:
  \begin{equation}
    \lim_{\lambda\to\infty} \mathfrak{J}_{\lambda,E}(\mu_{\mathrm{obs}}; \mu_\lambda) = \frac{1}{2} \delta_E^T \widetilde{\Sigma}_{EE}^{-1} \delta_E.
  \end{equation}
  The approximation residual satisfies $\mathfrak{J}_{\lambda,E} = \frac{1}{2} \delta_E^T \widetilde{\Sigma}_{EE}^{-1} \delta_E + O(\lambda^{-4})$.
\end{theorem}
\begin{proof}
  Substitute $t = x_E^*$ and $t - m_{\lambda,E} = -\lambda^{-1} \delta_E$ into \Cref{thm:multidim_cancellation}. Under the canonical scaling,
  \[
    \Sigma_{\lambda,EE}
    =
    \lambda^{-2}\widetilde\Sigma_{EE},
    \qquad
    t-m_{\lambda,E}
    =
    -\lambda^{-1}\delta_E,
    \qquad
    \|\rho_{\lambda,E}\|=O(\lambda^{-4}).
  \]
  Therefore the quadratic term yields exactly
  \[
    \frac12
    \delta_E^T
    \widetilde\Sigma_{EE}^{-1}
    \delta_E,
  \]
  and the determinant term satisfies $\log\det(I-\rho_{\lambda,E})=O(\lambda^{-4})$. Hence
  \[
    \mathfrak{J}_{\lambda,E}
    =
    \frac12
    \delta_E^T
    \widetilde\Sigma_{EE}^{-1}
    \delta_E
    +
    O(\lambda^{-4}).
  \]
\end{proof}

\subsection{Non-Gaussian and dynamical extensions}
\label{sec:discussion:non-gaussian}

The result is deliberately limited to finite-rank covariance block algebra and
its Gaussian Radon application.  It does not assert a general non-Gaussian
conditioning theory.  For non-Gaussian stationary processes, the projection
residual $S_{\lambda,E}=\Sigma_{\lambda,EE}-R_{\lambda,E}$ still has a
second-moment interpretation, but Radon disintegration, finite-part
functionals, and measure-class behavior require separate hypotheses
\cite{bogachev2007measure,kallenberg2021foundations}.  Dynamical relaxation
after conditioning, or transport distances between post-conditioning laws,
belongs to a different problem.  The contribution here is the
operator-theoretic removal of the compact-block Schur obstruction.

    % ============================================================
    %  \S7  CONCLUSION
    % ============================================================
    \section{Conclusion}
    \label{sec:conclusion}

    The paper resolves the Schur-complement obstruction created by compact
    trace-class free covariance blocks.  Since such blocks have no bounded
    ambient inverse, the residual is defined by Anderson--Trapp shorting on the
    Cameron--Martin range.  Galerkin convergence follows from nested
    variational suprema, not algebraic inversion.  The Gaussian conditioning
    statements then follow as consequences: disintegration supplies the
    exact-evidence law, and Fredholm determinant/trace identities cancel the
    finite-rank anomalies.  The paper does not treat infinite-rank observation
    operators or non-Gaussian conditioning theory.

    \section*{Declaration of Generative AI and AI-Assisted Technologies in the Manuscript Preparation Process}

    During the preparation of this work, the author(s) used Google DeepMind's Antigravity assistant in order to merge the LaTeX manuscripts, clean up cross-referencing tags, format mathematical notation, and edit the draft for style consistency. After using this tool/service, the author(s) reviewed and edited the content as needed and take(s) full responsibility for the content of the published article.

    % ============================================================
    %  REFERENCES
    % ============================================================
    \bibliographystyle{plain}
    \bibliography{paper1_ref}

    % ============================================================
    %  APPENDIX
    % ============================================================
    \appendix

    \input{paper1_appendix}
\end{document}
